---
output: github_document
editor_options: 
  chunk_output_type: console
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# rtables

[![Build Status](https://travis-ci.org/Roche/rtables.svg?branch=master)](https://travis-ci.org/Roche/rtables/)


## Reporting tables with R

The `rtables` R package was designed to create and display complex tables with R. The cells in an `rtable` may contain
any high-dimensional data structure which can then be displayed with cell-specific formatting instructions. Currently,
`rtables` can be outputted in `ascii` and `html`.

Note: we have completely refactored the `rtables` package which is officially released on CRAN in December 2020. With this significant change please familiarize yourself with the new framework by reading the package vignettes.

`rtables` is developed and copy written by `F. Hoffmann-La Roche` and it is released open source under Apache
License Version 2. 

`rtables` development is driven by the need to create regulatory ready tables for health authority review. Some of the key requirements for this undertaking are listed below:

* cell values and their visualization separate (i.e. no string based tables)
  - values need to be programmatically accessible in their non-rounded state for cross-checking
* multiple values displayed within a cell
* flexible tabulation framework
* flexible formatting (cell spans, rounding, alignment, etc.)
* multiple output formats (html, ascii, latex, pdf, xml)
* flexible pagination in both horizontal and vertical directions
* distinguish between name and label in the data structure to work with CDISC standards
* title, footnotes, cell cell/row/column references

Note that the current state of `rtables` does not fulfill all of those requirements, however, `rtables` is still under
active development and we are working on adding the missing features.


## Installation

`rtables` is now available on CRAN and you can install the latest released version with:

```r
install.packages("rtables")
```

or you can install the latest stable version directly from GitHub with:

```r
devtool::install_github("Roche/rtables")
```

To install a frozen pre-release version of `rtables` based on the new Layouting and Tabulation API as presented at
user!2020 and JSM2020 run the following command in `R`:

```{r, eval = FALSE}
devtools::install_github("roche/rtables", ref="v0.3.3")
```

To install the latest development version of the new test version of `rtables` run

```{r, eval = FALSE}
devtools::install_github("roche/rtables", ref = "gabe_tabletree_work")
```


## Usage

We first begin with a demographic table alike example and then show the creation of a more complex table.


```{r}
library(rtables)

lyt <- basic_table() %>%
  split_cols_by("ARM") %>%
  analyze(c("AGE", "BMRKR1", "BMRKR2"), function(x, ...) {
    if (is.numeric(x)) {
      in_rows(
        "Mean (sd)" = c(mean(x), sd(x)),
        "Median" = median(x),
        "Min - Max" = range(x),
        .formats = c("xx.xx (xx.xx)", "xx.xx", "xx.xx - xx.xx")
      )
    } else if (is.factor(x) || is.character(x)) {
      in_rows(.list = list_wrap_x(table)(x))
    } else {
      stop("type not supproted")
    }
  })

build_table(lyt, ex_adsl)
```


```{r}
library(rtables)
library(dplyr)

## for simplicity grab non-sparse subset
ADSL = ex_adsl %>% filter(RACE %in% levels(RACE)[1:3])

biomarker_ave = function(x, ...) {
     val = if(length(x) > 0) round(mean(x), 2) else "no data"
     in_rows(
        "Biomarker 1 (mean)" = rcell(val)
     )
}

basic_table() %>%
  split_cols_by("ARM") %>%
  split_cols_by("BMRKR2") %>%
  add_colcounts() %>%
  split_rows_by("RACE", split_fun = trim_levels_in_group("SEX")) %>%
  split_rows_by("SEX") %>%
  summarize_row_groups() %>%
  analyze("BMRKR1", biomarker_ave) %>%
  build_table(ADSL)
```


# Acknowledgements

We would like to thank everyone who has made `rtables` a better project by providing feedback and improving examples & vignettes. The following list of contributors is alphabetical:

Maximo Carreras, Francois Collins, Saibah Chohan, Tadeusz Lewandowski, Nick Paszty, Nina Qi, Jana Stoilova, Heng Wang, Godwin Yung

## Presentations

### New (Current) Layouting and Tabulation Framework (v.0.3+)

* [useR!2020 Presentation (on v0.3.1.1) July 2020](https://www.youtube.com/watch?v=CBQzZ8ZhXLA)

### v0.1.0 and previous

* [Presentation on v0.1.0 April 2018](https://docs.google.com/presentation/d/1bpdBDp4PZdZ4hCsfaPkAuHDVnJmtp7WBIZ19oKMDq0M/edit?usp=sharing)

* [baselR November
2017](https://docs.google.com/presentation/d/1V28AVo9aVNfw2FTuRgQyM4BJKalVQMFD8lKUD2KlzKI/edit?usp=sharing),
this presentation was written for version `v0.0.1`
