---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

# rtables

[![Build Status](https://travis-ci.org/Roche/rtables.svg?branch=master)](https://travis-ci.org/Roche/rtables/)


## Reporting tables with R

The `rtables` R package is a prototype to create and display complex tables with
R. The cells in an `rtable` may contain any high-dimensional data structure
which can then be displayed with cell-specific formatting instructions. Currently,
`rtables` can be outputted in `ascii` and `html`.

Some of our requirements and reasons to create this new table package are:

* multiple values displayed within a cell
* values need to be programmatically accessible for cross-checking
* flexible formatting (cell spans, rounding, alignment, etc.)
* composition of an `rtable` object from other rtable objects
* multiple output formats (html, ascii, latex, pdf)
* flexible pagination

Note that the current state of `rtables` does not fulfill all of those
requirements. We are looking for collaborators, see the next section.

## Collaboration and Planned API Changes

The `rtables` package is currently in the process of being reworked to address 

The current state of the `rtables` functionality is a first rough prototype. We
will be changing the API significantly in the coming months without maintaining
backward compatibility. We will release the `rtable` package on
[CRAN](https://cran.r-project.org/) once we have a stable version.

We are looking for collaborators to help define and shape the `rtables` package.
Please [contact me](mailto:adrian.waddell@roche.com) if you would like to align
your contributions to the current development plans or feel free to fork the
repository and to send us pull requests with the suggested improvements.


## Installation

To install a frozen pre-release version of `rtables` based on the new Layouting and Tabulation API as presented at user!2020 and JSM2020 run the following command in `R`:

```{r, eval = FALSE}
devtools::install_github("roche/rtables", ref="v0.3.2.3")
```

To install the latest development version of the new test version of `rtables` run

```{r, eval = FALSE}
devtools::install_github("roche/rtables", ref = "gabe_tabletree_work")
```


## Usage

```{r}
library(rtables)
library(dplyr)

## for simplicity grab non-sparse subset
ADSL = ex_adsl %>% filter(RACE %in% levels(RACE)[1:3])
biomarker_ave = function(x, ...) {
     val = if(length(x) > 0) round(mean(x), 2) else "no data"
     in_rows(
        "Biomarker 1 (mean)" = rcell(val)
     )
}

basic_table() %>%
  split_cols_by("ARM") %>%
  split_cols_by("BMRKR2") %>%
  add_colcounts() %>%
  split_rows_by("RACE", split_fun = trim_levels_in_group("SEX")) %>%
  split_rows_by("SEX") %>%
  summarize_row_groups() %>%
  analyze("BMRKR1", biomarker_ave) %>%
  build_table(ADSL)
```


# Acknowledgements

We would like to thank everyone who has made `rtables` a better project by providing feedback and improving examples & vignettes. The following list of contributors is alphabetical:

Daniel Sabanes Bove, Francois Collins, Tadeusz Lewandowski, Nick Paszty, Nina Qi, Jana Stoilova, Heng Wang.

## Presentations

### New (Current) Layouting and Tabulation Framework (v.0.3+)

* [useR!2020 Presentation (on v0.3.1.1) July 2020](https://www.youtube.com/watch?v=CBQzZ8ZhXLA)

### v0.1.0 and previous

* [Presentation on v0.1.0 April 2018](https://docs.google.com/presentation/d/1bpdBDp4PZdZ4hCsfaPkAuHDVnJmtp7WBIZ19oKMDq0M/edit?usp=sharing)

* [baselR November
2017](https://docs.google.com/presentation/d/1V28AVo9aVNfw2FTuRgQyM4BJKalVQMFD8lKUD2KlzKI/edit?usp=sharing),
this presentation was written for version `v0.0.1`
