---
title: "Comparing Against Baselines or Control"
author: "Gabriel Becker and Adrian Waddell"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Comparing Against Baselines or Control}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---



## Introduction


Often the data from one column is considered the reference/baseline/comparison group and is compared to the data from the other columns.

For example, lets calculate the average age:

```{r}
library(rtables)

basic_table() %>%
  split_cols_by("ARM") %>%
  analyze("AGE") %>%
  build_table(DM)
```

and then the difference of the average `AGE` between the arms to the placebo arm:

```{r}
basic_table() %>%
  split_cols_by("ARM", ref_group = "B: Placebo") %>%
  analyze("AGE", afun = function(x, .ref_group) {
    in_rows(
      "Difference of Averages" = rcell(mean(x) - mean(.ref_group), format = "xx.xx")
    )
  }) %>%
  build_table(DM)
```

Note that the column order has changed and the reference group is displayed in the first column.

In case we want to display a blank cell for "B: Placebo" we can test if a calculation happens for a cell that is part of the reference column with the `.in_ref_col` argument in `afun`:

```{r}
basic_table() %>%
  split_cols_by("ARM", ref_group = "B: Placebo") %>%
  analyze("AGE", afun = function(x, .ref_group, .in_ref_col){
    if (.in_ref_col) {
      in_rows("Difference of Averages" = rcell(NULL))
    } else {
      in_rows("Difference of Averages" = rcell(mean(x) - mean(.ref_group), format = "xx.xx"))
    }
  }) %>%
  build_table(DM)


basic_table() %>%
  split_cols_by("ARM", ref_group = "B: Placebo") %>%
  analyze("AGE", afun = function(x, .ref_group, .in_ref_col){
    if (.in_ref_col) {
      in_rows(tmp = NULL, tmp1 = NULL)
    } else {
      in_rows(
        "Difference of Averages" = rcell(mean(x) - mean(.ref_group), format = "xx.xx"),
         "another row" = rcell("aaa")
      )
    }
  }) %>%
  build_table(DM)

```

You can see which arguments are available for `afun` in the manual fro `analyze`.

## Row Splitting

When adding row-splitting the reference data might be one represented by the column with or without row splitting. For example:

```{r}
basic_table() %>%
  split_cols_by("ARM", ref_group = "B: Placebo") %>%
  add_colcounts() %>%
  split_rows_by("SEX", split_fun = drop_split_levels) %>%
  analyze("AGE", afun = function(x, .ref_group, .ref_full, .in_ref_col){
    in_rows(
      "is reference (.in_ref_col)" = rcell(.in_ref_col),
      "ref cell N (.ref_group)" = rcell(length(.ref_group)),
      "ref column N (.ref_full)" = rcell(length(.ref_full))
    )
  }) %>%
  build_table(subset(DM, SEX %in% c("M", "F")))
```

so the data assigned to `.ref_full` is the full data of the reference column where the data assigned to `.ref_group` respects the subsetting defined by row-splitting and hence is from the same subset as the argument `x` or `df` to `afun`.


