---
title: "Tables in Clinical Trials"
author: "Gabriel Becker and Adrian Waddell"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Tables in Clinical Trials}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, echo=FALSE}
knitr::opts_chunk$set(comment = "")
```


```{css, echo=FALSE}
.reveal .r code {
    white-space: pre;
}
```


## Introduction

In this vignette we create a demographic table, adverse event table, disposition table, and a response table using layouts. We don't aim to create fully specified tables that meet the requirements of most standard analyses, instead we only want to demonstrate how the layout based tabulation framework can specify the structure and relations that are commonly found when analyzing clinical trials data. At Roche we have another R package called `tern` to create standard tables with full parameterization to meet the requirements of the analysts.

Note that all the data is created using random number generators. All `ex_*` data which is currently attached to the `rtables` package were created using `random.cdisc.data` another R package that we intend to release as open source soon.

The packages used in this vignette are:

```{r, message=FALSE}
library(rtables)
library(tibble)
library(dplyr)
```

## Demographic Table

Demographic tables summarize variables for different population subsets (encoded in the columns).

One feature of `analyze` that we have not introduced yet is that the analysis function `afun` can specify multiple rows with the `in_rows` function:

```{r}
ADSL <- ex_adsl

basic_table() %>%
  split_cols_by("ARM") %>%
  analyze(vars = "AGE", afun = function(x) {
    in_rows(
      "Mean (sd)" = rcell(c(mean(x), sd(x)), format = "xx.xx (xx.xx)"),
      "Range" = rcell(range(x), format = "xx.xx - xx.xx")
    )
  }) %>%
  build_table(ADSL)
```

Multiple variables can be analyzed in one `analyze` call:

```{r}
basic_table() %>%
  split_cols_by("ARM") %>%
  analyze(vars = c("AGE", "BMRKR1"), afun = function(x) {
    in_rows(
      "Mean (sd)" = rcell(c(mean(x), sd(x)), format = "xx.xx (xx.xx)"),
      "Range" = rcell(range(x), format = "xx.xx - xx.xx")
    )
  }) %>%
  build_table(ADSL)
```

Hence, if `afun` can process different data vector types (i.e. variables selected from the data) then we are fairly close to a standard demographic table. Here is a function that creates a count table for factors and some summary calculation for numerical variables:


```{r}
s_summary <- function(x) {
  if (is.numeric(x)) {
    in_rows(
      "n" = rcell(sum(!is.na(x)), format = "xx"),
      "Mean (sd)" = rcell(c(mean(x, na.rm = TRUE), sd(x, na.rm = TRUE)), format = "xx.xx (xx.xx)"),
      "IQR" = rcell(IQR(x, na.rm = TRUE), format = "xx.xx"),
      "min - max" = rcell(range(x, na.rm = TRUE), format = "xx.xx - xx.xx")
    )
  } else if (is.factor(x)) {
    
    vs <- as.list(table(x))
    do.call(in_rows, lapply(vs, rcell, format = "xx"))
    
  } else (
    stop("type not supported")
  )
}
```

Note we use `rcell`s to wrap the results as it adds formatting instructions. We can use `s_summary` outside the context of tabulation:

```{r}
s_summary(ADSL$AGE)
```

and 

```{r}
s_summary(ADSL$SEX)
```

We can now create a commonly used demographic table:
 
```{r}
lyt <- basic_table() %>% 
  split_cols_by(var = "ARM") %>%
  analyze(c("AGE", "SEX"), afun = s_summary) 

tbl <- build_table(lyt, ADSL)
tbl
```

Note that `analyze` can also be called multiple times in sequence:

```{r}
tbl2 <- basic_table() %>% 
  split_cols_by(var = "ARM") %>%
  analyze("AGE", s_summary) %>%
  analyze("SEX", s_summary) %>%
  build_table(ADSL) 

tbl2
```

which leads to the identical table as `tbl1`:

```{r}
identical(tbl, tbl2)
```

```{r, echo=FALSE}
stopifnot(identical(tbl, tbl2))
```

In clinical trials analyses the number of patients per column is often referred to as `N` (rather than the overall population which is commonly referred to as `N`). Column `N`s are added using the `add_colcounts` function:

```{r}
basic_table() %>% 
  split_cols_by(var = "ARMCD") %>%
  add_colcounts() %>%
  analyze(c("AGE", "SEX"), s_summary) %>%
  build_table(ADSL) 
```

### Variations on the demographic table

We will now show a couple of variations of the demographic. These variations are in structure and not in analysis, hence they don't require a modification to the `s_summary` function.


We will start with a standard table analyzing the variables `AGE` and `BMRKR2` variables:

```{r}
basic_table() %>% 
  split_cols_by(var = "ARM") %>%
  add_colcounts() %>%
  analyze(c("AGE", "BMRKR2"), s_summary) %>%
  build_table(ADSL) 
```

let's assume we would like to have this analysis carried out per gender encoded in the row space:

```{r}
basic_table() %>% 
  split_cols_by(var = "ARM") %>%
  add_colcounts() %>%
  split_rows_by("SEX") %>%
  analyze(c("AGE", "BMRKR2"), s_summary) %>%
  build_table(ADSL) 
```

We will subset `ADSL` to include only males and females in the analysis in order to reduces the number of rows in the table


```{r}
ADSL_M_F <- filter(ADSL, SEX %in% c("M", "F"))

basic_table() %>% 
  split_cols_by(var = "ARM") %>%
  add_colcounts() %>%
  split_rows_by("SEX") %>%
  analyze(c("AGE", "BMRKR2"), s_summary) %>%
  build_table(ADSL_M_F) 
```

Note that the `UNDIFFERENTIATED` and `U` level still show up in the table. This is because tabulation respects the factor levels and level order, exactly as the `split` and `table` function. If empty levels should be dropped then the `split_*` function needs to know that via the  `split_fun` argument which excepts a splitting function, i.e. `drop_split_levels` here. Splitting is a bigger topic and we will address this in another vignette in more details. 
 
 
```{r}
basic_table() %>% 
  split_cols_by(var = "ARM") %>%
  add_colcounts() %>%
  split_rows_by("SEX", split_fun = drop_split_levels, child_labels = "visible") %>%
  analyze(c("AGE", "BMRKR2"), s_summary) %>%
  build_table(ADSL_M_F) 
```


Also, `M` and `F` are not very descriptive, you can add the full label with:

```{r}
ADSL_M_F_l <- ADSL_M_F %>% 
  mutate(lbl_sex = case_when(
    SEX == "M" ~ "Male",
    SEX == "F" ~ "Female",
    SEX == "U" ~ "Unknown",
    SEX == "UNDIFFERENTIATED" ~ "Undifferentiated"
  ))

basic_table() %>% 
  split_cols_by(var = "ARM") %>%
  add_colcounts() %>%
  split_rows_by("SEX", labels_var = "lbl_sex", split_fun = drop_split_levels, child_labels = "visible") %>%
  analyze(c("AGE", "BMRKR2"), s_summary) %>%
  build_table(ADSL_M_F_l)
```

**TODO: `labels_var` is not working**.

If we only want to stratify by gender for the `AGE` analyis one can do this using the `nested` argument:

```{r}
basic_table() %>% 
  split_cols_by(var = "ARM") %>%
  add_colcounts() %>%
  split_rows_by("SEX", split_fun = drop_split_levels, child_labels = "visible") %>%
  analyze("AGE", s_summary, show_labels = "visible") %>%
  analyze("BMRKR2", s_summary, nested = FALSE,  show_labels = "visible") %>%
  build_table(ADSL_M_F) 
```

Once we split the rows into groups (`M` and `F` here) one might want to summarize groups usually by showing count and column percentages. This is especially important if we have missing data. For example for:

```{r}
insert_NAs <- function(x) {
  x[sample(c(TRUE, FALSE), length(x), TRUE, prob = c(0.2, 0.8))] <- NA
  x
}

set.seed(1)
ADSL_NA <- ADSL %>% 
  mutate(AGE = insert_NAs(AGE))

basic_table() %>% 
  split_cols_by(var = "ARM") %>%
  add_colcounts() %>%
  split_rows_by("SEX", split_fun = drop_split_levels, child_labels = "visible") %>%
  analyze("AGE", s_summary) %>%
  analyze("BMRKR2", s_summary, nested = FALSE,  show_labels = "visible") %>%
  build_table(filter(ADSL_NA, SEX %in% c("M", "F"))) 
```

Here it is not easy to see how many females and males there are in each arm as `n` represents the number of non-missing data elements in the variables. Groups within rows that are defined by splitting can be summarized with `summarize_row_groups`, for example:

```{r}
basic_table() %>% 
  split_cols_by(var = "ARM") %>%
  add_colcounts() %>%
  split_rows_by("SEX", split_fun = drop_split_levels) %>%
  summarize_row_groups()  %>% 
  analyze("AGE", s_summary) %>%
  analyze("BMRKR2", afun = s_summary, nested = FALSE,  show_labels = "visible") %>%
  build_table(filter(ADSL_NA, SEX %in% c("M", "F"))) 
```

There are a couple of things to note here. Group summaries produce "content" rows. Visually it's impossible to distinguish data rows from content rows. Their difference is justified (and it's an important design decision) because when we paginate tables the content rows are by default repeated if a group gets divided via pagination. Conceptually the content rows summarize the patient population which is analyzed and hence is often the count & group percentages (default behavior of `summarize_row_groups`). We can recreate this default behavior (count percentage) by defining a `cfun` for illustrative purposes here as it results in the same table as above:

```{r}
basic_table() %>% 
  split_cols_by(var = "ARM") %>%
  add_colcounts() %>%
  split_rows_by("SEX", split_fun = drop_split_levels) %>%
  summarize_row_groups(cfun = function(df, labelstr, .N_col, ...) {
    rcell(nrow(df) * c(1, 1/.N_col), format = "xx (xx.xx%)", label = labelstr)
  })  %>% 
  analyze("AGE", s_summary) %>%
  analyze("BEP01FL", afun = s_summary, nested = FALSE,  show_labels = "visible") %>%
  build_table(filter(ADSL_NA, SEX %in% c("M", "F"))) 
```

Note that `cfun` differs from `afun` (which is used in `analyze`)  in that `cfun` does not operate on variables but rather on `data.frame`s or `tibble`s which are passed via the `df` argument (`afun` can optionally request `df` too). Further, `cfun` gives the default group label (factor level from splitting) as an argument to `labelstr` and hence it could be modified:

```{r}
basic_table() %>% 
  split_cols_by(var = "ARM") %>%
  split_rows_by("SEX", split_fun = drop_split_levels, child_labels = "hidden") %>%
  summarize_row_groups(cfun = function(df, labelstr, .N_col, ...) {
    rcell(nrow(df) * c(1, 1/.N_col), format = "xx (xx.xx%)", label = paste(labelstr, " count (perc.)"))
  })  %>% 
  analyze("AGE", s_summary) %>%
  analyze("BEP01FL", s_summary, nested = FALSE, show_labels = "visible") %>%
  build_table(filter(ADSL_NA, SEX %in% c("M", "F"))) 
```

**TODO: remove the `label` argument from `rcell` and specify the content row label with another mechanism**.  


### Using Layouts

Layouts have a couple of advantages over tabulating the tables directly:

* the creation of layouts requires the analyst to describe the problem in an abstract way
   - separate the analyses description from the actual data
* referencing variable names are via strings (no NSE needed, though this is arguably a feature or a short comming)
* layouts can be reused

Here is an example of the reusability of layouts:

```{r}
lyt <- NULL %>% 
  split_cols_by("ARM") %>%
  add_colcounts() %>%
  analyze(c("AGE", "SEX"), afun = s_summary)

lyt
```

We can now build a table for `ADSL`

```{r}
build_table(lyt, ADSL)
```

or for all patients that are older than 18:

```{r}
build_table(lyt, ADSL %>% filter(AGE > 18))
```


## Adverse Events

There are a number of different adverse event tables. We will now present two tables that show adverse events by id and then by grade and by id. 

This time we won't use the `ADAE` dataset from `random.cdisc.data` but rather generate on on the fly (see [Adrian's 2016 Phuse paper](https://github.com/waddella/phuse2016_adverse_events)):

```{r}
set.seed(1)

lookup <- tribble(
  ~AEDECOD,                          ~AEBODSYS,                                         ~AETOXGR,
  'HEADACHE',                        "NERVOUS SYSTEM DISORDERS",                        "5",
  'BACK PAIN',                       "MUSCULOSKELETAL AND CONNECTIVE TISSUE DISORDERS", "2",
  'GINGIVAL BLEEDING',               "GASTROINTESTINAL DISORDERS",                      "1",
  'HYPOTENSION',                     "VASCULAR DISORDERS",                              "3",
  'FAECES SOFT',                     "GASTROINTESTINAL DISORDERS",                      "2",
  'ABDOMINAL DISCOMFORT',            "GASTROINTESTINAL DISORDERS",                      "1",
  'DIARRHEA',                        "GASTROINTESTINAL DISORDERS",                      "1",
  'ABDOMINAL FULLNESS DUE TO GAS',   "GASTROINTESTINAL DISORDERS",                      "1",
  'NAUSEA (INTERMITTENT)',           "GASTROINTESTINAL DISORDERS",                      "2",
  'WEAKNESS',                        "MUSCULOSKELETAL AND CONNECTIVE TISSUE DISORDERS", "3",
  'ORTHOSTATIC HYPOTENSION',         "VASCULAR DISORDERS",                              "4"
)

normalize <- function(x) x/sum(x)
weightsA <- normalize(c(0.1, dlnorm(seq(0, 5, length.out = 25), meanlog = 3)))
weightsB <- normalize(c(0.2, dlnorm(seq(0, 5, length.out = 25))))

N_pop <- 300
ADSL2 <- data.frame(
  USUBJID = seq(1, N_pop, by = 1),
  ARM = sample(c('ARM A', 'ARM B'), N_pop, TRUE),
  SEX = sample(c('F', 'M'), N_pop, TRUE),
  AGE = 20 + rbinom(N_pop, size=40, prob=0.7)
)
                                      
l.adae <- mapply(ADSL2$USUBJID, ADSL2$ARM, ADSL2$SEX, ADSL2$AGE, FUN = function(id, arm, sex, age) {
  n_ae <- sample(0:25, 1, prob = if (arm == "ARM A") weightsA else weightsB)
  i <- sample(1:nrow(lookup), size = n_ae, replace = TRUE, prob = c(6, rep(1, 10))/16)
  lookup[i, ] %>% 
    mutate(
      AESEQ = seq_len(n()),
      USUBJID = id, ARM = arm, SEX = sex, AGE = age
    )
}, SIMPLIFY = FALSE)

ADAE2 <- do.call(rbind, l.adae)
ADAE2 <- ADAE2 %>% 
  mutate(
    ARM = factor(ARM, levels = c("ARM A", "ARM B")),
    AEDECOD = as.factor(AEDECOD),
    AEBODSYS = as.factor(AEBODSYS), 
    AETOXGR = factor(AETOXGR, levels = as.character(1:5))
  ) %>% 
  select(USUBJID, ARM, AGE, SEX, AESEQ, AEDECOD, AEBODSYS, AETOXGR)
  
ADAE2
```



### Adverse Events By ID

We start by defining an events summary function:

```{r}
s_events_patients <- function(x, labelstr, .N_col) {
  in_rows(
    "Total number of patients with at least one event" = 
      rcell(length(unique(x)) * c(1, 1/.N_col), format = "xx (xx.xx%)"),
    
    "Total number of events" = rcell(length(x), format = "xx")
  )
}
```

So, for a population `5` patients where one patient has 2 AEs, one patient has 1 AE, and 3 patients have no AEs we would get the following summary:

```{r}
s_events_patients(x = c("id 1", "id 1", "id 2"), .N_col = 5)
```

Let's use this summary function in a table:

```{r}
basic_table() %>% 
  split_cols_by("ARM") %>%
  add_colcounts() %>%
  analyze("USUBJID", s_events_patients) %>%
  build_table(ADAE2)
```

Note that the column `N`'s are wrong as by default they are set to the number of rows per group (i.e. number of AEs per arm here). This also affects the percentages. For this table we are interested by the number of patients per column/arm which is usually taken from `ADSL`:

```{r}
N_per_arm <- table(ADSL2$ARM)
N_per_arm
```

Since this information is not "pre-data" it needs to go to the table creation function `build_table`:

```{r}
basic_table() %>% 
  split_cols_by("ARM") %>%
  add_colcounts() %>%
  analyze("USUBJID", s_events_patients) %>%
  build_table(ADAE2, col_counts = N_per_arm)
```

We next calculate this information per system organ class:

```{r}
l <- basic_table() %>% 
  split_cols_by("ARM") %>%
  add_colcounts() %>%
  analyze("USUBJID", s_events_patients) %>%
  split_rows_by("AEBODSYS", child_labels = "visible", nested = FALSE)  %>%
  summarize_row_groups("USUBJID", cfun = s_events_patients)
  build_table(l, ADAE2, col_counts = N_per_arm)
```

We now have to the add a count table of `AEDECOD` for each `AEBODSYS`. The default `analyze` behavior for a factor is to create the count table per level (using `rtab_inner`):

```{r}
tbl1 <- basic_table() %>% 
  split_cols_by("ARM") %>%
  add_colcounts() %>%
  split_rows_by("AEBODSYS", child_labels = "visible")  %>%
  summarize_row_groups("USUBJID", cfun = s_events_patients) %>%
  analyze("AEDECOD", show_labels = "hidden", indent_mod = -1) %>%
  build_table(ADAE2, col_counts = N_per_arm)

tbl1
```

The `indent_mod` argument enables relative indenting changes if the tree structure of the table does not result in the desired indentation.


This table is however not per standard as it counts the total number of events and not the number of subjects one or more events for a particular term. To get the correct table we need to write a custom analysis function:

```{r}
table_count_once_per_id <- function(df, termvar = "AEDECOD", idvar = "USUBJID", na.rm = TRUE) {

  x <- df[[termvar]]
  id <- df[[idvar]]
  
  if (na.rm) x <- na.omit(x)
 
  as.list(table(x[!duplicated(id)]))
}

table_count_once_per_id(ADAE2)
```

So the desired AE table is:

```{r}
basic_table() %>%
  split_cols_by("ARM") %>%
  add_colcounts() %>%
  split_rows_by("AEBODSYS", child_labels = "visible")  %>%
  summarize_row_groups("USUBJID", cfun = s_events_patients) %>%
  analyze("AEDECOD", afun = table_count_once_per_id, show_labels = "hidden", indent_mod = -1) %>%
  build_table(ADAE2, col_counts = N_per_arm)
```

**TODO: afun should accept x and df**

Note that we are missing the overall summary. This can be added with another `analyze` call and then setting `nested = FALSE` in the subsequent `summarize_row_groups` call (**TODO: though that is not intuitive we will likely change that** as it does not make sense in the layout framwork to nest a split in an analyisis and currently the `nested` argument needs to be explicit):

```{r}
basic_table() %>% 
  split_cols_by("ARM") %>%
  add_colcounts() %>%
  analyze("USUBJID", afun = s_events_patients) %>% 
  split_rows_by("AEBODSYS", child_labels = "visible", nested = FALSE)  %>%
  summarize_row_groups("USUBJID", cfun = s_events_patients) %>%
  analyze("AEDECOD", table_count_once_per_id, show_labels = "hidden", indent_mod = -1) %>%
  build_table(ADAE2, col_counts = N_per_arm)
```


### Adverse Events By ID and By Grade

For this table we do not show the zero count grades. Note that we add overall groups with a custom split function. 

```{r}
table_count_grade_once_per_id <- function(df, labelstr = "", gradevar = "AETOXGR", idvar = "USUBJID", grade_levels = NULL, na.rm = TRUE) {
  
  id <- df[[idvar]]
  grade <- df[[gradevar]]
  
  if (!is.null(grade_levels)) {
    stopifnot(all(grade %in% grade_levels))
    grade <- factor(grade, levels = grade_levels)
  }
  
  if (na.rm) grade <- na.omit(grade)
  
  id_sel <- !duplicated(id)
  
  in_rows(
      "--Any Grade--" = sum(id_sel),
      .list =  as.list(table(grade[id_sel]))
    )
}

table_count_grade_once_per_id(ex_adae, grade_levels = 1:5)
```


```{r}
basic_table() %>% 
  split_cols_by("ARM") %>%
  add_colcounts() %>%
  analyze("AETOXGR", 
          afun = table_count_grade_once_per_id, 
          extra_args = list(grade_levels = 1:5),
          var_labels = "- Any adverse events -", show_labels = "visible") %>%
  split_rows_by("AEBODSYS", child_labels = "visible", nested = FALSE) %>%
  summarize_row_groups(cfun = table_count_grade_once_per_id, format = "xx", indent_mod = 1) %>%
  split_rows_by("AEDECOD", child_labels = "visible", indent_mod = -2)  %>%
  analyze("AETOXGR", 
          afun = table_count_grade_once_per_id, 
          extra_args = list(grade_levels = 1:5), show_labels = "hidden") %>%
  build_table(ADAE2)
```

** TODO: now the overall label is missing ***


## Disposition Table

Let's start with creating a suitable `ADSL` (currently not part of `random.cdisc.data`):

```{r}
set.seed(1)
ADSL0 <- ex_adsl %>%
  mutate(
    COMPSTUD = sample(c('Y','N'), size=nrow(ex_adsl), replace = TRUE) %>% as.factor,
    STUDONS = sample(c('Alive: On Treatment', 'Alive: In Follow-up', NA), size = nrow(ex_adsl), replace = TRUE) %>% as.factor,
    STDDRS = sample(c('Death', 'Lost To Follow-Up',
                      'Protocol Violation', 'Withdrawal By Subject',
                      'Other'),
                    size=nrow(ex_adsl),
                    replace = TRUE) %>% as.factor,
    GOTTRT = ifelse(!is.na(ACTARMCD), 'Y', 'N') %>% as.factor,
    DISTRTFL = sample(c('Y','N'),
                      size=nrow(ex_adsl),
                      replace = TRUE) %>% as.factor,
    TRTDRS = sample(c('ADVERSE EVENT', 'PROGRESSIVE DISEASE',
                      'PHYSICIAN DECISION', 'LACK OF EFFICACY',
                      'OTHER'),
                    size=nrow(ex_adsl),
                    replace = TRUE) %>% as.factor,
    STUDONS = case_when(COMPSTUD == 'N' ~ STUDONS),
    STDDRS = case_when(COMPSTUD == 'N' & is.na(STUDONS) ~ STDDRS),
    DISSTDFL = case_when(!is.na(STDDRS) ~ "Y"),
    DISTRTFL = case_when(GOTTRT == 'Y' ~ DISTRTFL),
    TRTDRS = case_when(DISTRTFL == 'Y' ~ TRTDRS),
    DRSCAT = case_when(
      TRTDRS %in% c('ADVERSE EVENT', 'PHYSICIAN DECISION') ~ "Safety",
      !is.na(TRTDRS) ~ "Other"
    )
  ) %>% var_relabel(COMPSTUD = "Complete Study",
                    STUDONS = "On-study Status",
                    DISSTDFL = "Discontinued Study",
                    STDDRS = "Reason for Study \r\nDiscontinuation",
                    GOTTRT = "Received Treatment",
                    DISTRTFL = "Discontinued Treatment",
                    TRTDRS = "Reason for Treatment \r\nDiscontinuation",
                    DRSCAT = "Subcategory for Treatment Discontinuation"
  )

ADSL0
```

We also need a couple of utility functions to deal with the particular format:

```{r}
yn_as_bool <- function(f) {
  function(x, .N_col) {
    f(x == "Y", .N_col)
  }
}

s_count_perc <- function(x, .N_col) {
  rcell(sum(!is.na(x)) * c(1, 1/.N_col), format = "xx (xx.xx%)")
}
```

A dispositon table is then created as follows:

```{r disposition_table}
basic_table() %>%
  split_cols_by("ARM") %>%
  analyze("COMPSTUD", yn_as_bool(s_count_perc), var_labels = "Completed Study") %>%
  split_rows_by("DISSTDFL", nested = FALSE, split_fun = keep_split_levels("Y")) %>%
  summarize_row_groups(label_fstr = "Discontinued Study") %>%
  split_rows_by("DRSCAT", split_fun = reorder_split_levels(c("Safety", "Other"))) %>%
  summarize_row_groups() %>%
  analyze("TRTDRS", var_labels = "Discontinued Study") %>% 
  build_table(ADSL0) %>%
  trim_zero_rows()
```

## Response Table

The response table that we will create here is composed of 3 parts:

1. Binary response table
1. Unstratified analysis comparison vs. control group
1. Multinomial response table


Let's start with the first part which is fairly simple to derive:

```{r}
ADRS_BESRSPI <- ex_adrs %>%
  filter(PARAMCD == "BESRSPI") %>%
  mutate(
    rsp = factor(AVALC %in% c("CR", "PR"), levels = c(TRUE, FALSE), labels = c("Responders", "Non-Responders")),
    is_rsp = (rsp == "Responders")
  )

s_proportion <- function(x, .N_col) {
   in_rows(.list = lapply(as.list(table(x)), function(xi) rcell(xi * c(1, 1/.N_col), format = "xx.xx (xx.xx%)")))
}

basic_table() %>%
  split_cols_by("ARMCD", ref_group = "ARM A") %>%
  add_colcounts() %>%
  analyze("rsp", s_proportion, show_labels = "hidden") %>%
  build_table(ADRS_BESRSPI)
```


Note that we have set the `ref_group` argument in `split_cols_by` which for the current table has no effect as we only use the cell data for the responder and non-responder counting.

We will now look at 2. Unstratified analysis comparison vs. control group. Let's start with the analysis function:

```{r, eval = FALSE}
s_unstratified_response_analysis <- function(x, .ref_group, .in_ref_col) {
  
  if (.in_ref_col) {
    return(in_rows(
        "Difference in Response Rates (%)" = rcell(numeric(0)),
        "95% CI (Wald, with correction)" = rcell(numeric(0)),
        "p-value (Chi-Squared Test)" = rcell(numeric(0)),
        "Odds Ratio (95% CI)" = rcell(numeric(0))
    ))
  }
  
  fit <- stats::prop.test(
    x = c(sum(x), sum(.ref_group)),
    n = c(length(x), length(.ref_group)),
    correct = FALSE
  )
  
  fit_glm <- stats::glm(
    formula = rsp ~ group,
    data = data.frame(
      rsp = c(.ref_group, x), 
      group = factor(rep(c("ref", "x"), times = c(length(.ref_group), length(x))), levels = c("ref", "x"))
    ),
    family = binomial(link = "logit")
  )

  in_rows(
    "Difference in Response Rates (%)" = rcell((mean(x) - mean(.ref_group)) * 100, format = "xx.xx") ,
    "95% CI (Wald, with correction)" = rcell(fit$conf.int * 100, format = "(xx.xx, xx.xx)"),
    "p-value (Chi-Squared Test)" = rcell(fit$p.value, format = "x.xxxx | (<0.0001)"),
    "Odds Ratio (95% CI)" = rcell(c(
      exp(stats::coef(fit_glm)[-1]),
      exp(stats::confint.default(fit_glm, level = .95)[-1, , drop = FALSE])
    ), format = "xx.xx (xx.xx - xx.xx)")
  )
}

s_unstratified_response_analysis(
  x = ADRS_BESRSPI %>% filter(ARM == "A: Drug X") %>% pull(is_rsp), 
  .ref_group = ADRS_BESRSPI %>% filter(ARM == "B: Placebo") %>% pull(is_rsp),
  .in_ref_col = FALSE
)
```

so hence we can now add the next section to the table:

```{r, eval = FALSE}
basic_table() %>%
  split_cols_by("ARMCD", ref_group = "ARM A") %>%
  add_colcounts() %>%
  analyze("rsp", s_proportion, show_labels = "hidden") %>%
  analyze("is_rsp", s_unstratified_response_analysis) %>%
  build_table(ADRS_BESRSPI)
```

** TODO: discuss indenting**

And next we will add the Multinomial response table. We are doing a row-split by response level, and then doing the same thing as we did for the binary response table above. However, when we want to do row-split, that means we need to also take into account the `.N_col` as total and then reconstruct the full logical vector.

```{r, eval = FALSE}
s_prop <- function(df, .N_col) {
  in_rows(
    "95% CI (Wald, with correction)" = rcell(binom.test(nrow(df), .N_col)$conf.int * 100, format = "(xx.xx, xx.xx)")
  )
}

s_prop(
  df = ADRS_BESRSPI %>% filter(ARM == "A: Drug X", AVALC == "CR"), 
  .N_col = sum(ADRS_BESRSPI$ARM == "A: Drug X")
)
```

Note that we have added the `labelstr` argument **TODO: this is clunky**

```{r, eval = FALSE}
basic_table() %>%
  split_cols_by("ARMCD", ref_group = "ARM A") %>%
  add_colcounts() %>%
  analyze("rsp", s_proportion, show_labels = "hidden") %>%
  analyze("is_rsp", s_unstratified_response_analysis) %>%
  split_rows_by(
    var = "AVALC",
    split_fun = reorder_split_levels(neworder = c("CR", "PR", "SD", "NON CR/PD", "PD", "NE"), drlevels = TRUE), 
    nested = FALSE
  ) %>%
  summarize_row_groups() %>%
  analyze("AVALC", afun = s_prop) %>%
  build_table(ADRS_BESRSPI)
```

In case the we wanted to rename the levels of `AVALC` and remove the CI for `NE` we could do that as follows:


```{r, eval = FALSE}
rsp_label <- function(x) {
  rsp_full_label <- c(
    CR          = "Complete Response (CR)",
    PR          = "Partial Response (PR)",
    SD          = "Stable Disease (SD)",
    `NON CR/PD` = "Non-CR or Non-PD (NON CR/PD)",
    PD          = "Progressive Disease (PD)",
    NE          = "Not Evaluable (NE)",
    Missing     = "Missing",
    `NE/Missing` = "Missing or unevaluable"
  )
  stopifnot(all(x %in% names(rsp_full_label)))
  rsp_full_label[x]
}


tbl <- basic_table() %>%
  split_cols_by("ARMCD", ref_group = "ARM A") %>%
  add_colcounts() %>%
  analyze("rsp", s_proportion, show_labels = "hidden") %>%
  analyze("is_rsp", s_unstratified_response_analysis) %>%
  split_rows_by(
    var = "AVALC",
    split_fun = keep_split_levels(c("CR", "PR", "SD", "NON CR/PD", "PD"), reorder = TRUE), 
    nested = FALSE
  ) %>%
  summarize_row_groups(cfun = function(df, labelstr, .N_col) {
    in_rows(nrow(df) * c(1, 1/.N_col), .labels = rsp_label(labelstr))
  }) %>%
  analyze("AVALC", afun = s_prop) %>%
  analyze("AVALC", afun = function(x, .N_col) {
    in_rows(rcell(sum(x == "NE") * c(1, 1/.N_col), format = "xx.xx (xx.xx%)"), .labels = rsp_label("NE"))
  }, nested = FALSE) %>%
  build_table(ADRS_BESRSPI)

tbl
```

