---
title: "Column Counts and Formats"
author: "Davide Garolini"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introspecting Tables}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  chunk_output_type: console
---

```{r, echo=FALSE}
knitr::opts_chunk$set(comment = "#")
```

# Column Counts and Formats

Main drivers of column counts are parameters `show_colcounts` and `colcount_format` in `split_cols_by()`. The former is a boolean that determines whether to show the column counts in the table, while the latter is a string that specifies the format of the column counts. The default value for `show_colcounts` is `FALSE`, while the default value for `colcount_format` is `"(N=xx)"`.

```{r}
library(dplyr)
library(rtables)

example_data <- data.frame(
  ARMCD = c("ARM A", "ARM A", "ARM A", "ARM A", "ARM B", "ARM B"),
  STRATA2 = c("S1", "S2", "S1", "S2",  "S2", "S2"),
  STRATA1 = c("A", "A", "B", "B", "A", "B"),
  AGE = c(20, 30, 40, 50, 21, 24)
)

lyt <- basic_table() %>%
  split_cols_by("ARMCD", show_colcounts = TRUE, colcount_format = "N=xx") %>%
  split_cols_by("STRATA2", show_colcounts = TRUE) %>%
  split_cols_by("STRATA1") %>%
  add_overall_col("All") %>%
  analyze("AGE", afun = max, format = "xx.x")

tbl <- build_table(lyt, example_data)
tbl
```

We can control the column counts with the following functions:

```{r}
col_counts(tbl) # vector of all of them (visible and not visible) -> difficult to read/use for nested structures
coldf <- make_col_df(tbl) # to check the structure of the columns as a data.frame
coltree_structure(tbl) # Prints the structure of the columns as a tree
```

`make_col_df` is a fundamental function that is used also internally to keep track of the tree structure of the columns. It returns a `data.frame` very similar to `make_row_df`, but for columns. The `coltree_structure` function is a wrapper around `make_col_df` that prints the structure of the columns as a tree.

To retrieve the column counts for a specific column, use pathing! Also visibility can be accessed and modified with the following functions:

```{r}
facet_colcount(tbl, coldf$path[[1]]) # To get the column counts for a specific column at a specific level
colcount_visible(tbl, coldf$path[[1]]) # not visible!!! # FALSE
facet_colcount(tbl, coldf$path[[1]][c(1, 2)]) # visible
colcount_visible(tbl, coldf$path[[1]][c(1, 2)]) # TRUE
```

Formats can differ as we have shown in the example above. We can change the format of the column counts with the following function:

```{r, eval = FALSE}
# Modifying last level!?
colcount_format(tbl) <- "[N<=xx]" # Does not work
colcount_visible(tbl, coldf$path[[1]]) <- TRUE # not the right way!!
# > tbl # ERROR! we need to set the other values too
# Error in h(simpleError(msg, call)) :
#   error in evaluating the argument 'x' in selecting a method for function 'toString':
#   Detected different colcount visibility among sibling facets (those arising from the
#   same split_cols_by* layout instruction). This is not supported.
# Set count values to NA if you want a blank space to appear as the displayed count for particular facets.
# First disagreement occured at paths:
# ARMCD[ARM A]->STRATA2[S1]->STRATA1[A]
# ARMCD[ARM A]->STRATA2[S1]->STRATA1[B]
# colcount_visible(tbl, coldf$path[[1]]) <- FALSE

# Trying with col_counts
col_counts(tbl) <- c(1, rep(NA_integer_, length(col_counts(tbl)) - 1)
# colcount_visible(tbl, coldf$path[[1]]) <- TRUE # same error as above we need facets!!
facet_colcounts_visible(tbl, c("ARMCD", "*", "STRATA2", "*", "STRATA1")) <- TRUE # does not work
print(tbl)
```

It is also possible to set the column counts and visibility for a specific column level with pathing and the following function:

```{r}
# Modifying another level/column
facet_colcount(tbl, c("ARMCD", "ARM A")) <- NA_integer_ # To remove the column counts for a specific column
# facet_colcount(tbl, "ARMCD") # not working
print(tbl)

# colcount_visible(tbl, c("*")) <- TRUE # it does not work!! We need facet_colcount_visible
facet_colcounts_visible(tbl, c("ARMCD")) <- FALSE  # function needs examples and connection to colcounts_visible
print(tbl)
```

A general rule is that we can set the column counts for a single facet or for multiple facets. The following table summarizes the functions that can be used for this purpose:

```
               single facet     | multiple facets
_______________________________________________________
count         facet_colcount    | *
visibility    colcount_visible  | facet_colcounts_visible          

```
*: There is no general "set multiple counts simultaneously" function, though we do support `col_counts<-` for setting all the leaf counts at the same time for backwards compatibility.

Formats can be chosen from `formatters::list_valid_format_labels()` for one dimension or two with `%` (at the moment is only heritage, it will always be `100%`). It is also possible to use a function for formats.

```{r, eval = FALSE}
tbl2 <- basic_table() %>%
  split_cols_by("ARMCD", show_colcounts = TRUE, colcount_format = "N=xx (xx%)") %>%
  split_cols_by("STRATA2", show_colcounts = TRUE, colcount_format = function(x, ...) {browser()}) %>%
  analyze("AGE") %>%
  build_table(example_data)
tbl2
```

For removing the column counts from a specific column, we can do it in two ways:

```{r}
ccnts <- col_counts(tbl)
ccnts[1] <- NA_integer_
col_counts(tbl) <- ccnts
tbl
```

It is possible also to change (for now uniformly only) the output string in case of missing values in the column counts. The default value is `""`, that will show nothing. We can change it with the following function:

```{r}
colcount_na_str(tbl) <- "NaN"
facet_colcount(tbl, coldf$path[[1]][c(1, 2)]) <- NA_integer_
tbl
```

Finally, we can remove all column counts from a table with the following function:

```{r}
rm_all_colcounts(tbl) # returns the table without column counts
```
