---
title: "Customizing Appearance"
author: "Davide Garolini, Abinaya Yogasekaram, and Gabriel Becker"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Customizing Appearance}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, echo=FALSE}
knitr::opts_chunk$set(comment = "#")
```

```{css, echo=FALSE}
.reveal .r code {
    white-space: pre;
}
```

## Customizing Appearance

In this vignette, we describe the various ways we can modify and customize the appearance of `rtables`.

Loading the package:

```{r, message=FALSE}
library(rtables)
```

### Rows and cell values alignments

It is possible to align the content by assigning `"left"`, `"center"` (default), and `"right"` to `.aligns` and `align` arguments in `in_rows()` and `rcell()`, respectively.

Please consider using `?in_rows` and `?rcell` for further clarifications on the two arguments, and use `rtables_aligns()` to see all available alignment options.

In the following we show two simplified examples that use `align` and `.aligns`, respectively.

```{r}
# In rcell we use align.
lyt <- basic_table() %>%
  analyze("AGE", function(x) {
    in_rows(
      left = rcell("l", align = "left"),
      right = rcell("r", align = "right"),
      center = rcell("c", align = "center")
    )
  })

tbl <- build_table(lyt, DM)
tbl
```

```{r}
# In in_rows, we use .aligns. This can either set the general value or the
#   single values (see NB).
lyt2 <- basic_table() %>%
  analyze("AGE", function(x) {
    in_rows(
      left = rcell("l"),
      right = rcell("r"),
      center = rcell("c"),
      .aligns = c("right")
    ) # NB: .aligns = c("right", "left", "center")
  })

tbl2 <- build_table(lyt2, DM)
tbl2
```

These concepts can be well applied to any clinical table as shown in the following, more complex, example.

```{r}
lyt3 <- basic_table() %>%
  split_cols_by("ARM") %>%
  split_rows_by("SEX") %>%
  analyze(c("AGE", "STRATA1"), function(x) {
    if (is.numeric(x)) {
      in_rows(
        "mean" = rcell(mean(x)),
        "sd" = rcell(sd(x)),
        .formats = c("xx.x"), .aligns = "left"
      )
    } else if (is.factor(x)) {
      rcell(length(unique(x)), align = "right")
    } else {
      stop("Unsupported type")
    }
  }, show_labels = "visible", na_str = "NE")

tbl3 <- build_table(lyt3, ex_adsl)
tbl3
```

### Top-left Materials

The sequence of strings printed in the area between the column header display and the first row label \emph{top left material} can be modified during pre-processing using label position argument in row splits `split_rows_by`, with the `append_topleft` function,  and during post-processing using the `top_left()` function. Note: Indenting is automatically added \emph{only in the case of} `label_pos = "topleft"`.

Within the layout initializer:

```{r}
lyt <- basic_table() %>%
  split_cols_by("ARM") %>%
  split_rows_by("STRATA1") %>%
  analyze("AGE") %>%
  append_topleft("New top_left material here")

build_table(lyt, DM)
```

Specify label position using the `split_rows` function. Notice the position of `STRATA1` and `SEX`.

```{r}
lyt <- basic_table() %>%
  split_cols_by("ARM") %>%
  split_rows_by("STRATA1", label_pos = "topleft") %>%
  split_rows_by("SEX", label_pos = "topleft") %>%
  analyze("AGE")

build_table(lyt, DM)
```

Post-processing using the `top_left()` function:

```{r}
lyt <- basic_table() %>%
  split_cols_by("ARM") %>%
  split_rows_by("SEX") %>%
  analyze(c("AGE", "STRATA1"), function(x) {
    if (is.numeric(x)) {
      in_rows(
        "mean" = rcell(mean(x)),
        "sd" = rcell(sd(x)),
        .formats = c("xx.x"), .aligns = "left"
      )
    } else if (is.factor(x)) {
      rcell(length(unique(x)), align = "right")
    } else {
      stop("Unsupported type")
    }
  }, show_labels = "visible", na_str = "NE") %>%
  build_table(ex_adsl)

# Adding top-left material
top_left(lyt) <- "New top-left material here"

lyt
```

### Table Inset

Table title, table body, referential footnotes and and main footers can be inset from the left alignment of the titles and provenance footer materials. This can be modified within the layout initializer `basic_table()` using the `inset` argument or during post-processing with `table_inset()`.

Using the layout initializer:

```{r}
lyt <- basic_table(inset = 5) %>%
  analyze("AGE")

build_table(lyt, DM)
```

Using the post-processing function:

Without inset - 
```{r}
lyt <- basic_table() %>%
  analyze("AGE")

tbl <- build_table(lyt, DM)
tbl
```

With an inset of 5 characters - 

```{r}
table_inset(tbl) <- 5
tbl
```


Below is an example with a table produced for clinical data. Compare the inset of the table and main footer between the two tables.

Without inset -
```{r}
analysisfun <- function(x, ...) {
    in_rows(row1 = 5,
            row2 = c(1, 2),
            .row_footnotes = list(row1 = "row 1 rfn"),
            .cell_footnotes = list(row2 = "row 2 cfn"))
    }

lyt <- basic_table(title = "Title says Whaaaat", subtitles = "Oh, ok.",
                   main_footer = "ha HA! Footer!", prov_footer = "provenaaaaance") %>%
    split_cols_by("ARM") %>%
    analyze("AGE", afun = analysisfun)

result <-  build_table(lyt, ex_adsl)
result
```

With inset - 
Notice, the inset does not apply to any title materials (main title, subtitles, page titles), or provenance footer materials. Inset settings is applied to top-left materials, referential footnotes main footer materials and any horizontal dividers.

```{r}
table_inset(result) <- 5
result
```

### Horizontal Separation

A character value can be specified to modify the horizontal separation between column headers and the table. Horizontal separation applies when:\
1) separating title + subtitles from the column labels + topleft materials,\
2) column labels + top left material from row labels + cells,\
3) row labels + cells from footer content, and\
4) Referential footnotes from main + provenance content \emph{only if} there would be something on both sides of the divider.\

Below, we replace the default line with "=".

```{r}
tbl <- basic_table() %>%
  split_cols_by("Species") %>%
  add_colcounts() %>%
  analyze(c("Sepal.Length", "Petal.Width"), function(x) {
    in_rows(
      mean_sd = c(mean(x), sd(x)),
      var = var(x),
      min_max = range(x),
      .formats = c("xx.xx (xx.xx)", "xx.xxx", "xx.x - xx.x"),
      .labels = c("Mean (sd)", "Variance", "Min - Max")
    )
  }) %>%
  build_table(iris, hsep = "=")
tbl
```


### Section Dividers

A character value can be specified as a section divider which succeed every group defined by a split instruction. Note, a trailing divider at the end of the table is never printed.

Below, a "+" is repeated and used as a section divider.

```{r}
lyt <- basic_table() %>%
  split_cols_by("Species") %>%
  analyze(head(names(iris), -1), afun = function(x) {
    list(
      "mean / sd" = rcell(c(mean(x), sd(x)), format = "xx.xx (xx.xx)"),
      "range" = rcell(diff(range(x)), format = "xx.xx")
    )
  }, section_div = "+")

build_table(lyt, iris)
```

Section dividers can be set to " " to create a blank line.

```{r}
lyt <- basic_table() %>%
  split_cols_by("Species") %>%
  analyze(head(names(iris), -1), afun = function(x) {
    list(
      "mean / sd" = rcell(c(mean(x), sd(x)), format = "xx.xx (xx.xx)"),
      "range" = rcell(diff(range(x)), format = "xx.xx")
    )
  }, section_div = " ")

build_table(lyt, iris)
```

Separation characters can be specified for different row splits. However, only one will be printed if they "pile up" next to each other. 

```{r}
lyt <- basic_table() %>%
  split_cols_by("ARM") %>%
  split_rows_by("RACE", section_div = "=") %>%
  split_rows_by("STRATA1", section_div = "~") %>%
  analyze("AGE", mean, var_labels = "Age", format = "xx.xx")

build_table(lyt, DM)
```

### Indent Modifier

Tables by default have indenting at each level of splitting. A custom indent value can be supplied with the `indent_mod` argument within a split function to modify this default. Compare the indenting of the tables below:

Default Indent -

```{r}
basic_table(
  title = "Study XXXXXXXX",
  subtitles = c("subtitle YYYYYYYYYY", "subtitle2 ZZZZZZZZZ"),
  main_footer = "Analysis was done using cool methods that are correct",
  prov_footer = "file: /path/to/stuff/that/lives/there HASH:1ac41b242a"
) %>%
  split_cols_by("ARM") %>%
  split_rows_by("SEX") %>%
  split_rows_by("STRATA1") %>%
  analyze("AGE", mean, format = "xx.x") %>%
  build_table(DM)
```

Modified indent - 
```{r}
basic_table(
  title = "Study XXXXXXXX",
  subtitles = c("subtitle YYYYYYYYYY", "subtitle2 ZZZZZZZZZ"),
  main_footer = "Analysis was done using cool methods that are correct",
  prov_footer = "file: /path/to/stuff/that/lives/there HASH:1ac41b242a"
) %>%
  split_cols_by("ARM") %>%
  split_rows_by("SEX", indent_mod = 3) %>%
  split_rows_by("STRATA1", indent_mod = 5) %>%
  analyze("AGE", mean, format = "xx.x") %>%
  build_table(DM)
```

### Variable Label Visibility

With split instructions, visibility of the label for the variable being split can be modified to `visible`, `hidden` and `topleft` with the `show_labels` argument, `label_pos` argument, and `child_labels` argument where applicable. Note: this is NOT the name of the levels contained in the variable. For analyze calls, \code{default} indicates that the variable should be visible only if multiple variables are analyzed at the same level of nesting.

Visibility of labels for the groups generated by a split can also be modified using the `child_label` argument with a split call. The `child_label` argument can force labels to be visible in addition to content rows but we cannot hide or move the content rows.

Notice the placement of the "AGE" label in this example:

```{r}
lyt <- basic_table(show_colcounts = TRUE) %>%
  split_cols_by(var = "ARM") %>%
  split_rows_by("SEX", split_fun = drop_split_levels, child_labels = "visible") %>%
  split_rows_by("STRATA1") %>%
  analyze("AGE", mean, show_labels = "default")

build_table(lyt, DM) 
```
When set to default, the label `AGE` is not repeated since there is only one variable being analyzed at the same level of nesting. Override this by setting the `show_labels` argument as "visible". 

```{r}
lyt2 <- basic_table(show_colcounts = TRUE) %>%
  split_cols_by(var = "ARM") %>%
  split_rows_by("SEX", split_fun = drop_split_levels, child_labels = "hidden") %>%
  split_rows_by("STRATA1") %>%
  analyze("AGE", mean, show_labels = "visible")

build_table(lyt2, DM) 
```

Below is an example using the `label_pos` argument for modifying label visibility:

Label order will mirror the order of `split_rows_by` calls. If the labels of any subgroups should be hidden, the `label_pos` arugment should be set to hidden. 

"SEX" label position is hidden - 
```{r}
basic_table(
  title = "Study XXXXXXXX",
  subtitles = c("subtitle YYYYYYYYYY", "subtitle2 ZZZZZZZZZ"),
  main_footer = "Analysis was done using cool methods that are correct",
  prov_footer = "file: /path/to/stuff/that/lives/there HASH:1ac41b242a"
) %>%
  split_cols_by("ARM") %>%
  split_rows_by("SEX", split_fun = drop_split_levels, label_pos = "visible") %>%
  split_rows_by("STRATA1", label_pos = "hidden") %>%
  analyze("AGE", mean, format = "xx.x") %>%
  build_table(DM)
```

"SEX" label position is with the top-left materials - 

```{r}
basic_table(
  title = "Study XXXXXXXX",
  subtitles = c("subtitle YYYYYYYYYY", "subtitle2 ZZZZZZZZZ"),
  main_footer = "Analysis was done using cool methods that are correct",
  prov_footer = "file: /path/to/stuff/that/lives/there HASH:1ac41b242a"
) %>%
  split_cols_by("ARM") %>%
  split_rows_by("SEX", split_fun = drop_split_levels, label_pos = "topleft") %>%
  split_rows_by("STRATA1", label_pos = "hidden") %>%
  analyze("AGE", mean, format = "xx.x") %>%
  build_table(DM)
```