---
title: "Customizing Appearance"
author: "Davide Garolini and Gabriel Becker"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Customizing Appearance}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---


```{r, echo=FALSE}
knitr::opts_chunk$set(comment = "#")
```

```{css, echo=FALSE}
.reveal .r code {
    white-space: pre;
}
```

## Customizing Appearance


### Rows and cell values alignments

It is possible to align the content by assigning `"left"`, `"center"` (default), and `"right"` to `.aligns` and `align` arguments in `in_rows()` and `rcell()`, respectively.

Please consider using `?in_rows` and `?rcell` for further clarifications on the two arguments, and use `rtables_aligns()` to see all available alignment options.

In the following we show two simplified examples that use `align` and `.aligns`, respectively.

```{r}
# In rcell we use align.
lyt <- basic_table() %>%
    analyze("AGE", function(x) {
        in_rows(left = rcell("l", align = "left"),
                right = rcell("r", align = "right"),
                center = rcell("c", align = "center"))
    })

build_table(lyt, DM)
```

```{r}
# In in_rows, we use .aligns. This can either set the general value or the
#   single values (see NB).
lyt2 <- basic_table() %>%
    analyze("AGE", function(x) {
        in_rows(left = rcell("l"),
                right = rcell("r"),
                center = rcell("c"),
                .aligns = c("right")) # NB: .aligns = c("right", "left", "center")
    })

build_table(lyt2, DM)
```

These concepts can be well applied to any clinical table as shown in the following, more complex, example.

```{r}
basic_table() %>%
  split_cols_by("ARM") %>%
  split_rows_by("SEX") %>%
  analyze(c("AGE", "STRATA1"), function(x) {
    if (is.numeric(x)) {
      in_rows(
        "mean" = rcell(mean(x)),
        "sd" = rcell(sd(x)),
        .formats = c("xx.x"), .aligns = "left"
      )
    } else if (is.factor(x)) {
      rcell(length(unique(x)), align = "right")
    } else {
      stop("Unsupported type")
    }
  }, show_labels = "visible", na_str = "NE") %>%
  build_table(ex_adsl)
```
