---
title: "The Split Machinery"
author: "Davide Garolini"
date: '`r Sys.Date()`'
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rtables Advanced Usage}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Disclaimer

This vignette is currently under development. Any code or prose which
appears in a version of this vignette on the `main` branch of the
repository will work/be correct, but they likely are not in their
final form.


## The Split Machinery

The scope of this vignette is understanding how `rtables` creates facets by splitting the incoming data into hierarchical groups that go from the root node to singular `rcell`s. The latter level, also called leaf-level, contains the final partition that is subjected to the analysis functions. More details from the user perspective can be found in relevant vignette `split_functions` and function documentation like `?split_rows_by` and `?split_funcs`.

The following will describe how the split machinery mainly for the row domain. Further information on how columns are defined will follow.


## `do_split`

```{r, message=FALSE}
library(rtables)
```
