---
title: "Formats Precedence and NA Handling"
author: "Wojciech Wójciak and Gabriel Becker"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Formats Precedence NA Handling}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, echo=FALSE}
knitr::opts_chunk$set(comment = "#")
```

```{css, echo=FALSE}
.reveal .r code {
  white-space: pre;
}
```

## Formats precedence

A user of `rtables` package can specify the format in which the numbers in the reporting tables are printed. Formatting functionally is provided by the [`formatters`](https://insightsengineering.github.io/formatters/) R package and available formats can be listed with `formatters::list_valid_format_labels()` function.
The format can be specified by the user in a few different places. It may happen that for a single table layout, the format is specified in more than one place. In such a case, the final format that will be applied, depends on formats precedence rules used in `rtables`. In this vignette, we describe basic rules of `rtables` formats precedence.

The packages used in this vignette are:

```{r, message=FALSE}
library(rtables)
```

The examples shown in this vignette are based on the example `ADSL` dataset, i.e. a demographic table that summarizes the variables content for different population subsets (encoded in the columns).

```{r}
ADSL <- ex_adsl
```

Note that all `ex_*` data which is currently attached to the `rtables` package is provided by the [`formatters`](https://insightsengineering.github.io/formatters/) package and was created using the publicly available [`random.cdisc.data`](https://insightsengineering.github.io/random.cdisc.data/) R package.

### Basic formats precedence and inheritance rules

The format in which numbers are printed can be specified by the user in a few different places. In the context of the precedence, it is important at which level of the split hierarchy these formats are specified.  . In general, there are two such levels: the cell level, row level or so called parent table level. 

TODO cell <- parent_table <- parent_table ... <- parent_table

### No format explicitly specified.

Below is a simple layout that does not explicitly set any specific format for the output of the analysis function used in this example. In such a case, the format from function `base::as.character()` will be used.

```{r}
lyt0 <- basic_table() %>%
  split_cols_by("ARM") %>%
  analyze(vars = "AGE", afun = mean)

build_table(lyt0, ADSL)
```

### Cell format

The format of the cell can be explicitly specified through `rcell()` or `in_rows()` functions. The former is essentially a collection of data objects while the latter is a collection of `rcell()` objects. In the context of inheritance, this is the most specific place where the format can be specified by the user.

```{r}
lyt1 <- basic_table() %>%
  split_cols_by("ARM") %>%
  analyze(vars = "AGE", afun = function(x) {
    rcell(mean(x), format = "xx.xx", label = "Mean")
  })

build_table(lyt1, ADSL)

lyt1a <- basic_table() %>%
  split_cols_by("ARM") %>%
  analyze(vars = "AGE", afun = function(x) {
    in_rows(
      "Mean" = rcell(mean(x)),
      .formats = "xx.xx"
    )
  })

build_table(lyt1a, ADSL)
```

If the format is specified in both of these places at the same time, the one specified through `in_rows()`, has a higher precedence. Technically, in this case, the format defined in `rcell()` will simply be overwritten by the one defined in `in_rows()`.

```{r}
lyt2 <- basic_table() %>%
  split_cols_by("ARM") %>%
  analyze(vars = "AGE", afun = function(x) {
    in_rows(
      "Mean" = rcell(mean(x), format = "xx.xxx"),
      .formats = "xx.xx"
    )
  })

build_table(lyt2, ADSL)
```

### Parent table format and inheritance.

In addition to the cell level, the format can be specified at the parent table level(s). Then, if no format has been set by the user for a cell, the format from the nearest parent split table will be inherited.

```{r}
lyt3 <- basic_table() %>%
  split_cols_by("ARM") %>%
  analyze(vars = "AGE", mean, format = "xx.x")

build_table(lyt3, ADSL)
```

If the cell format was specified too, then the parent table format is ignored for this cell, since the cell format is treated as the most specific.

```{r}
lyt4 <- basic_table() %>%
  split_cols_by("ARM") %>%
  analyze(
    vars = "AGE", afun = function(x) {
      rcell(mean(x), format = "xx.xx", label = "Mean")
    },
    format = "xx.x"
  )

build_table(lyt4, ADSL)

lyt4a <- basic_table() %>%
  split_cols_by("ARM") %>%
  analyze(
    vars = "AGE", afun = function(x) {
      in_rows(
        "Mean" = rcell(mean(x)),
        "SD" = rcell(sd(x)),
        .formats = "xx.xx"
      )
    },
    format = "xx.x"
  )

build_table(lyt4a, ADSL)
```

In the following, slightly more complicated example, we can observe partial inheritance. That is, only `SD` cell inherits the parent table's format, while the `Mean` cell does not.

```{r}
lyt5 <- basic_table() %>%
  split_cols_by("ARM") %>%
  analyze(
    vars = "AGE", afun = function(x) {
      in_rows(
        "Mean" = rcell(mean(x), format = "xx.xx"),
        "SD" = rcell(sd(x))
      )
    },
    format = "xx.x"
  )

build_table(lyt5, ADSL)
```

### NA handling

TODO
