---
title: "Creating your own format values and labels"
author: "Martin Cadek"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating your own format values and labels}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r, echo=FALSE}
knitr::opts_chunk$set(comment = "#")
```

```{css, echo=FALSE}
.reveal .r code {
  white-space: pre;
}
```

```{r, message=FALSE}
library(rtables)
library(dplyr)
ADSL <- ex_adsl %>%
    mutate(ARM = factor(ARM), SEX = factor(SEX))
```

### Introduction

The  section will showcase how to extend the standard examples of `list_valid_format_labels()` to create your own format functions.  It is recommended that before this tutorial, the reader will familiarise themselves with the basic functionality of `formatters` package in the following [getting started page](https://insightsengineering.github.io/formatters/latest-tag/articles/formatters.html) for `formatters` package and the [introduction](https://insightsengineering.github.io/rtables/main/articles/introduction.html) to `rtables`. The next step is to understand the basics of the [format precedence](https://insightsengineering.github.io/rtables/main/articles/format_precedence.html#parent-table-format-and-inheritance).

We will use the following `ADSL` data frame to demo the different approaches. 

```{r adsl_analyzed, echo=FALSE}
lyt_no_format <- basic_table() %>%
  split_cols_by("ARM") %>%
  split_rows_by("SEX") %>%
  analyze(vars = "AGE", afun = mean)

adsl_analyzed <- build_table(lyt_no_format, ADSL)

adsl_analyzed
```


### Creating your own static format functions

When you run `list_valid_format_labels()`, you will see several pre-defined valid format labels for 1d, 2d, and 3d tables. 

```{r, echo=FALSE}
list_valid_format_labels()
```

This should give you a good idea what is possible. However, you may want to create your own format functions.