---
title: "Formatting Tables"
author: "Gabriel Becker and Adrian Waddell"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Formatting Tables}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

## Introduction


```{r, message=FALSE}
library(rtables)
library(dplyr)
```

## Cell Formats


Cell formats are inherited from `rtable` to `rrow` to `rcell` if there is not a
more specific format argument (i.e. format arguments of `rcell` are more
specific than those of `rrow` which are more specific than those of `rtable`.

Two types of formats are supported:

* format label, e.g. `xx.xx (xx.xx%)`

* functions that take the element and output type as an argument and return a 
formatted string or html object **note, we will eventually deprecate this**

Formats labels are currently limited to the following predefined set:

```{r}
list_rcell_format_labels()
```

Here is an example of using functions for formatting cells:

```{r}
my_format <- function(x, output) {
   paste(x, collapse = "/")
}
tbl3 <- rtable(
  c("A", "B"),
  format = my_format,
  rrow("row1", c(1,2,3,4), letters[1:10])
)
tbl3
```

## Output Types

### ASCII

#### PDF

#### Column Width / Row Heights

### HTML

#### Viewer

The `Viewer` functions takes an `rtable` object and displays it with
[bootstrap](https://getbootstrap.com/) styling in RStudio's Viewer pane.

```{r, eval = FALSE}
tbl <- basic_table() %>%
  split_cols_by("Species") %>%
  analyze("Sepal.Length", mean) %>%
  build_table(iris)

Viewer(tbl)
```

It is also possible to show two tables side by side

```{r, eval = FALSE}
tbl2 <- basic_table() %>%
  split_cols_by("Species") %>%
  analyze("Sepal.Width", mean) %>%
  build_table(iris)

Viewer(tbl, tbl2)
```

### LaTeX

### RTF

### Power Point










