---
title: "Creating your own format values and labels"
author: "Martin Cadek"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating your own format values and labels}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
  markdown: 
    wrap: 72
---

```{r, echo=FALSE}
knitr::opts_chunk$set(comment = "#")
```

```{css, echo=FALSE}
.reveal .r code {
  white-space: pre;
}
```

```{r, message=FALSE}
library(rtables)
library(dplyr)
ADSL <- ex_adsl %>%
  mutate(ARM = factor(ARM), SEX = factor(SEX))
```

### Introduction

The  section will showcase how to extend the standard examples of `list_valid_format_labels()` to create your own format functions.  It is recommended that before this tutorial, the reader will familiarise themselves with the basic functionality of `formatters` package in the following [getting started page](https://insightsengineering.github.io/formatters/latest-tag/articles/formatters.html) for `formatters` package and the [introduction](https://insightsengineering.github.io/rtables/main/articles/introduction.html) to `rtables`. The next step is to understand the basics of the [format precedence](https://insightsengineering.github.io/rtables/main/articles/format_precedence.html#parent-table-format-and-inheritance).

We will use the following layout to demo the different approaches. The layout is missing the `analyze` step so the output does not have generate table with any analysed variables, but we will add it later.

```{r adsl_analyzed, echo=FALSE}
basic_layout <- basic_table() %>%
  split_cols_by("ARM") %>%
  split_rows_by("SEX")

build_table(basic_layout, ADSL)
```

If we need to apply any standard formatting, we can simply look for one of the pre-defined formats from `list_valid_format_labels()`.

```{r, echo=FALSE}
list_valid_format_labels()
```

There are standard formats for 1d, 2d, and 3d tables. Applying standard formatting is done as part of adding analysed variables with `analyze` function. Specifically, it is part of the `format` argument. Let's add `"xx.xxx"` to format the mean of AGE.

```{r, echo=FALSE}
basic_layout %>%
  analyze(vars = "AGE", afun = mean, format = "xx.xxx") %>%
  build_table(ADSL)
```

Let's assume we want to create new format, however, that format is not part of the `list_valid_format_labels()`. The following sections cover various situation.

### Creating custom formats

Let's suppose we want to create a new format that will display three decimal places and also says "years" after the number. One way to do this is to tell the analyse function to paste a character vector containing a formatted combination of text and variable value. This is essentially what `sprintf` does which is a base R function.

For example, if you wish to format the value of `pi` to three decimal places, you can use `sprintf` as follows:
```{r, echo=FALSE}
sprintf("%0.3f", pi)
```

The `sprintf` is quite powerful and fast function, so it is quite helpful to read the documentation. In general, we could create many different formats using `sprintf`. Let's get creative and do some formatting below.

```{r, echo=FALSE}
# ... with 2 decimal places
sprintf("%.2f", pi)
# ... use width of 5 and leading zero
sprintf("%05.2f", pi)
# ... use width of 5 and leading space
sprintf("%5.2f", pi)
# ... scientific notation with 2 decimal places
sprintf("%e", pi)
# ... hexadecimal floating-point notation
sprintf("%a", pi)
# ... percentage with 2 decimal places
sprintf("%.2f%%", pi * 100)
# ... use width of 8, 3 decimal places, and leading zeros
sprintf("%08.3f", pi)
# ... with custom text
sprintf("The value of pi is approximately %.3f", pi)
```

How to use sprintf in `analyze` function? We would need to create a function that will use `sprintf` and then use this function in the `analyze` function. However, when we will pass this into the `analyze` function, we need to ensure that it will apply to all of the values in the our table. Notice that the examples above are for a single value. There are various ways to do this, one way is to use `lapply` function. Let's create a function that will use `sprintf` and `lapply` to format the values.

```{r, echo=FALSE}
sprintf_age_years <- function(x, ...) {
  lapply(x, function(item) sprintf("%0.3f years", item))
}

basic_layout %>%
  analyze(vars = "AGE", afun = mean, format = sprintf_age_years) %>%
  build_table(ADSL)
```

This is important for understanding the basis of customised formats; however, the `formatters` package provides an easier way to do this.

#### Using `sprintf_format` function

The function `sprintf_format` from `formatters` works as a wrapper around `sprintf`. This means that we do not need to implement `sprintf` inside the `lapply` function ourselves. The `sprintf_format` function will do something like this for us. Let's see how to use it.

```{r, echo=FALSE}
format_age_years <- sprintf_format("%0.3f years")
```

```{r, echo=FALSE}
basic_layout %>%
  analyze(vars = "AGE", afun = mean, format = format_age_years) %>%
  build_table(ADSL)
```

We can also use all of the `sprintf` formats shown in the previous section in the `sprintf_format` function. This is because the ineternals use `sprintf` function. For example, scientific notation with 2 decimal places:

```{r, echo=FALSE}
format_age_years_scipen <- sprintf_format("%e years")
```

```{r, echo=FALSE}
basic_layout %>%
  analyze(vars = "AGE", afun = mean, format = format_age_years_scipen) %>%
  build_table(ADSL)
```

#### Using `sprintf_format` on 2d and 3d tables
Let's continue with more complex examples. As seen below, we use the `sprintf_format` function on 2d and 3d tables. 


```{r, echo=FALSE}
format_mean_median_sd <- sprintf_format("%.2f - %.1f - %.0f")
```

We can use this format in the `analyze` function. Below is the example for table with values in mean, median, and sd. The format above is defined for three values separated by a dash.

```{r, echo=FALSE}
basic_layout %>%
  analyze(vars = "AGE", afun = function(x) {
    list(
      "Age: mean, median, sd" = rcell(c(mean(x), median(x) ,sd(x)))
    )
    }, format = format_mean_median_sd) %>%
  build_table(ADSL)
```

Extending the example to 3d table, we can use the same format. However, we need to add a few more format definitions to support all of the values in the table.

```{r, echo=FALSE}
format_mean_median_sd <- sprintf_format("%.2f - %.1f - %.0f")
format_max_min <- sprintf_format("%.0f - %.0f")
format_range <- sprintf_format("%.0f")
```


In this case, the most clear is to define the format for each values in the list separately. We can do this becauase the `rcell` function allows us to define the format.
```{r, echo=FALSE}
basic_layout %>%
  analyze(vars = "AGE", afun = function(x) {
    list(
      "Age: mean, median, sd" = rcell(c(mean(x), median(x), sd(x)), format = format_mean_median_sd),
      "Age: max, min" = rcell(c(max(x), min(x)), format = format_max_min),
      "Age: range" = rcell(max(x) - min(x), format = format_range)
    )
  }) %>%
  build_table(ADSL)
```

#### Using conditional custom formatting
Importantly, it is also possible to program completely custom formatting functions. 

Let's investigate what `format_value` function does. 

```{r, echo=FALSE}
format_value(3.14159265359, "xx.xxx")
```

When value is supplied, the function will return the value formatted according to the format string. If the value is missing, the function will return NA. If the function is ran on table, it will return the table with formatted values.

```{r, echo=FALSE}
basic_layout %>%
  analyze(vars = "AGE", afun = mean, format = format_value("xx.xx")) %>%
  build_table(ADSL)
```

How could we create our own custom format function? 

```{r, echo=FALSE}
custom_format_value <- function(x, format = NULL, na_str = "NA") {
  if (length(x) == 0) {
    return("")
  }
  
  if (length(na_str) == 0) {
    na_str <- "NA"
  }
  
  if (is.null(format)) {
    return(toString(x))
  }
  
  # Simplified switch statement for format handling
  switch(format,
    "xx" = as.character(x),
    "xx." = round(x, digits = 0),
    "xx.x" = round(x, digits = 1),
    "xx.xx" = round(x, digits = 2),
    "xx.xxx" = round(x, digits = 3),
    "xx.xxxx" = round(x, digits = 4),
    paste("format string", format, "not found")
  )
}

# Example usage
basic_layout %>%
  analyze(vars = "AGE", afun = mean, format = format_value("xx.xx")) %>%
  build_table(ADSL)



```



```{r, echo=FALSE}
list_valid_format_labels()
```

This should give you a good idea what is possible. However, you may want to create your own format functions.