---
title: "Formats Precedence and NA Handling"
author: "Wojciech Wójciak and Gabriel Becker"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Formats Precedence NA Handling}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, echo=FALSE}
knitr::opts_chunk$set(comment = "#")
```

```{css, echo=FALSE}
.reveal .r code {
  white-space: pre;
}
```

## Formats Precedence

A user of `rtables` package can specify the format in which the numbers in the reporting tables are printed. Formatting functionally is provided by the [`formatters`](https://insightsengineering.github.io/formatters/) R package and available formats can be listed with `formatters::list_valid_format_labels()` function.
The format can be specified by the user in a few different places. It may happen that for a single table layout, the format is specified in more than one place. In such a case, the final format that will be applied depends on formats precedence rules defined by `rtables`. In this vignette, we describe the basic rules of `rtables` formats precedence.

The packages used in this vignette are:

```{r, message=FALSE}
library(rtables)
```

The examples shown in this vignette are based on the example `ADSL` dataset, i.e. a demographic table that summarizes the variables content for different population subsets (encoded in the columns).

```{r}
ADSL <- ex_adsl
```

Note that all `ex_*` data which is currently attached to the `rtables` package is provided by the [`formatters`](https://insightsengineering.github.io/formatters/) package and was created using the publicly available [`random.cdisc.data`](https://insightsengineering.github.io/random.cdisc.data/) R package.

### Formats precedence and inheritance rules

The format in which numbers are printed can be specified by the user in a few different places. In the context of the precedence, it is important at which level of the split hierarchy formats are specified. In general, there are two such levels: the **cell** level or the so-called **parent table** level. The concept of the cell and the parent table results from the way in which `rtables` package stores resulting tables. It models the resulting tables as hierarchical, tree-like objects with the cells (as leaves) containing multiple values. Particularly noteworthy in this context is the fact that the actual table splitting occurs in a row-dominant way (even if the column split is present in the layout). `rtables` provides user-end function `table_structure()` that prints the structure of a given table object. For a quick illustration, consider the following example.

```{r}
lyt <- basic_table() %>%
  split_cols_by("ARM") %>%
  split_rows_by("SEX") %>%
  analyze(vars = "AGE", afun = mean)

adsl_analyzed <- build_table(lyt, ADSL)
adsl_analyzed
table_structure(adsl_analyzed)
```

For this reporting table, there are 4 sub-tables under `SEX` table. These are: `F`, `M`, `U`, and `UNDIFFERENTIATED`. Each of these sub-tables has one sub-table `AGE`. So, e.g. for the top sub-table `AGE`, its nearest parent table is `F`.

The concept of hierarchical, tree-like representation of resulting tables, translates directly to formats precedence and inheritance rules. As a general principle, the format being finally applied for the cell is the one that is the most specific, that is, the one which is the closest to that cell in a given path in the tree. Hence, the precedence-inheritance chain looks like below.
```
cell <- parent_table <- parent_table <- ... <- parent_table
```
In such a chain, the most outer `parent_table` is the least specific place to specify the format, while the `cell` is the most specific one. In case the format is specified by the user in more than one place, the one which is more specific will be applied for the cell. If no specific format has been selected by the user for the split, then the default format will be applied. The default format is "xx" and it yields the same formatting as `as.character()` function. In the following sections of this vignette, we will illustrate the format precedence rules with a few examples.

### Standard format

Below is a simple layout that does not explicitly set any specific format for the output of the analysis function used in this example. In such a case, the default format will be used.

```{r}
lyt0 <- basic_table() %>%
  split_cols_by("ARM") %>%
  analyze(vars = "AGE", afun = mean)

build_table(lyt0, ADSL)
```

### Cell format

The format of the cell can be explicitly specified through `rcell()` or `in_rows()` functions. The former is essentially a collection of data objects while the latter is a collection of `rcell()` objects. As already mentioned, this is the most specific place where the format can be specified by the user.

```{r}
lyt1 <- basic_table() %>%
  split_cols_by("ARM") %>%
  analyze(vars = "AGE", afun = function(x) {
    rcell(mean(x), format = "xx.xx", label = "Mean")
  })

build_table(lyt1, ADSL)

lyt1a <- basic_table() %>%
  split_cols_by("ARM") %>%
  analyze(vars = "AGE", afun = function(x) {
    in_rows(
      "Mean" = rcell(mean(x)),
      .formats = "xx.xx"
    )
  })

build_table(lyt1a, ADSL)
```

If the format is specified in both of these places at the same time, the one specified through `in_rows()`, has higher precedence. Technically, in this case, the format defined in `rcell()` will simply be overwritten by the one defined in `in_rows()`. This is because the format specified in `in_rows()` is applied to the cells not the rows (overriding the previously specified cell specific values), which means that the precedence rules described above are still in place.

```{r}
lyt2 <- basic_table() %>%
  split_cols_by("ARM") %>%
  analyze(vars = "AGE", afun = function(x) {
    in_rows(
      "Mean" = rcell(mean(x), format = "xx.xxx"),
      .formats = "xx.xx"
    )
  })

build_table(lyt2, ADSL)
```

### Parent table format and inheritance

In addition to the cell level, the format can be specified at the parent table level. Then, if no format has been set by the user for a cell, the most specific format for that cell is the one defined at the nearest parent split table (if any).

```{r}
lyt3 <- basic_table() %>%
  split_cols_by("ARM") %>%
  analyze(vars = "AGE", mean, format = "xx.x")

build_table(lyt3, ADSL)
```

If the cell format was specified too, then the parent table format is ignored for this cell, since the cell format is treated as more specific.

```{r}
lyt4 <- basic_table() %>%
  split_cols_by("ARM") %>%
  analyze(
    vars = "AGE", afun = function(x) {
      rcell(mean(x), format = "xx.xx", label = "Mean")
    },
    format = "xx.x"
  )

build_table(lyt4, ADSL)

lyt4a <- basic_table() %>%
  split_cols_by("ARM") %>%
  analyze(
    vars = "AGE", afun = function(x) {
      in_rows(
        "Mean" = rcell(mean(x)),
        "SD" = rcell(sd(x)),
        .formats = "xx.xx"
      )
    },
    format = "xx.x"
  )

build_table(lyt4a, ADSL)
```

In the following, slightly more complicated example, we can observe partial inheritance. That is, only `SD` cell inherits the parent table's format, while the `Mean` cell does not.

```{r}
lyt5 <- basic_table() %>%
  split_cols_by("ARM") %>%
  analyze(
    vars = "AGE", afun = function(x) {
      in_rows(
        "Mean" = rcell(mean(x), format = "xx.xx"),
        "SD" = rcell(sd(x))
      )
    },
    format = "xx.x"
  )

build_table(lyt5, ADSL)
```

## `NA` Handling

Consider the following layout and the resulting table created:

```{r}
lyt6 <- basic_table() %>%
  split_cols_by("ARM") %>%
  split_rows_by("SEX") %>%
  analyze(vars = "AGE", afun = mean, format = "xx.xx")

build_table(lyt6, ADSL)
```

In the output, the cell corresponding to `UNDIFFERENTIATED` level of `SEX` and `B: Placebo drug` level of `ARM` categorical variables is displayed as `NA`. This happened because there were no non-`NA` values under this facet to compute the mean. `rtables` allows the user to specify a string to display when cell values are `NA`. Similar to formats for numbers, the user can specify a string to replace `NA` with the parameter `format_na_str` or `.format_na_str`. This can be specified at the **cell** or **parent table** level. The `NA` string precedence and inheritance rules are the same as those for number formats precedence, described in the previous section of this vignette. We will illustrate them with a few examples.

### Replacing `NA` values at the cell level

At the cell level, it is possible to replace `NA` values with a custom string by means of the `format_na_str` parameter in `rcell()` or `.format_na_str` parameter in `in_rows()`.

```{r}
lyt7 <- basic_table() %>%
  split_cols_by("ARM") %>%
  split_rows_by("SEX") %>%
  analyze(vars = "AGE", afun = function(x) {
    rcell(mean(x), format = "xx.xx", label = "Mean", format_na_str = "<missing>")
  })

build_table(lyt7, ADSL)

lyt7a <- basic_table() %>%
  split_cols_by("ARM") %>%
  split_rows_by("SEX") %>%
  analyze(vars = "AGE", afun = function(x) {
    in_rows(
      "Mean" = rcell(mean(x), format = "xx.xx"),
      .format_na_strs = "<MISSING>"
    )
  })

build_table(lyt7a, ADSL)
```

If the `NA` string is specified in both of these places at the same time, the one specified with `in_rows()`, has higher precedence. Technically, in this case, the `NA` replacement string defined in `rcell()` will simply be overwritten by the one defined in `in_rows()`. This is because the `NA` string specified in `in_rows()` is applied to the cells not the rows (overriding the previously specified cell specific values), which means that the precedence rules described above are still in place.

```{r}
lyt8 <- basic_table() %>%
  split_cols_by("ARM") %>%
  split_rows_by("SEX") %>%
  analyze(vars = "AGE", afun = function(x) {
    in_rows(
      "Mean" = rcell(mean(x), format = "xx.xx", format_na_str = "<missing>"),
      .format_na_strs = "<MISSING>"
    )
  })

build_table(lyt8, ADSL)
```

### Parent table replacement of `NA` values and inheritance principles

In addition to the cell level, the string replacement for `NA` values can be specified at the parent table level. Then, if no replacement string has been specified by the user for a cell, the most specific `NA` string for that cell is the one defined at the nearest parent split table (if any).

```{r}
lyt9 <- basic_table() %>%
  split_cols_by("ARM") %>%
  split_rows_by("SEX") %>%
  analyze(vars = "AGE", mean, format = "xx.xx", na_str = "not available")

build_table(lyt9, ADSL)
```

If the replacement string for `NA` values was specified also at the cell level, then the one set at the parent table level is ignored for this cell, as the cell level settings are treated as the most specific.

```{r}
lyt10 <- basic_table() %>%
  split_cols_by("ARM") %>%
  split_rows_by("SEX") %>%
  analyze(
    vars = "AGE", afun = function(x) {
      rcell(mean(x), format = "xx.xx", label = "Mean", format_na_str = "<missing>")
    },
    na_str = "not available"
  )

build_table(lyt10, ADSL)

lyt10a <- basic_table() %>%
  split_cols_by("ARM") %>%
  split_rows_by("SEX") %>%
  analyze(
    vars = "AGE", afun = function(x) {
      in_rows(
        "Mean" = rcell(mean(x)),
        "SD" = rcell(sd(x)),
        .formats = "xx.xx",
        .format_na_strs = "<missing>"
      )
    },
    na_str = "not available"
  )

build_table(lyt10a, ADSL)
```

In the following, slightly more complicated example, we can observe partial inheritance. That is, only `SD` cell inherits the parent table's `NA` string, while the `Mean` cell does not.

```{r}
lyt11 <- basic_table() %>%
  split_cols_by("ARM") %>%
  split_rows_by("SEX") %>%
  analyze(
    vars = "AGE", afun = function(x) {
      in_rows(
        "Mean" = rcell(mean(x), format_na_str = "<missing>"),
        "SD" = rcell(sd(x))
      )
    },
    format = "xx.xx",
    na_str = "not available"
  )

build_table(lyt11, ADSL)
```
