---
title: "Advanced rtables - Structure-Conditional Behavior In `afun`s With `.spl_context`"
subtitle: Contributed by Johnson & Johnson Innovative Medicine
date: "2025-10-22"
author:
- Gabriel Becker
- Dan Hofstaedter
output:
    rmarkdown::html_document:
        theme: "spacelab"
        highlight: "kate"
        toc: true
        toc_float: true
        code_folding: show
vignette: >
  %\VignetteIndexEntry{Advanced rtables - Structure-Conditional Behavior In afuns With .spl_context}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
      wrap: 72
  chunk_output_type: console
---

```{r, include = FALSE}
suggested_dependent_pkgs <- c("dplyr")
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = all(vapply(
    suggested_dependent_pkgs,
    requireNamespace,
    logical(1),
    quietly = TRUE
  ))
)
```

```{r, echo=FALSE}
knitr::opts_chunk$set(comment = "#")
```

# Split Context

The split context (i.e., the optional `.spl_context` argument to a/c
funs) provides analysis functions the ability to know what
substructure of the table it is calculating cell values for, the data
(sub)set corresponding to that substructure, and the steps - both in
terms of faceting structure and corresponding data subsetting - taken
by the tabulation engine to arrive where it is now.

This allows us to program custom analysis functions which have
behavior *conditional on which row or column facet they are currently
calculating cell values for*, as we will see further down in this
document.

The *split context* is a `data.frame` with one row per faceting step
in row space up to and including the one the analysis funciton is
being called within, and the following columns which vary across
context rows:

- **`split`** (`character`) - name of the split represented by each
  row of the split context
- **`value`** (`character`) - string representation of the value of
  the split for each row of the context
- **`full_parent_df`** (`list` of `data.frame`s) - the full data
  (across all columns) corresponding to each row faceting step
- **`all_cols_n`** (`integer`) - the observation count for each row
  faceting step (across all columns)
- **`<column names>`** (one `list` column per column in the table) -
  logical vectors corresponding to the subset of `full_parent_df` for
  the named column for each faceting step. Named by
  `names(col_exprs(tab))`.

In addition, the context contains the following columns which are
constant across context rows:

- **`cur_col_id`** (`character`) - identifier for the current column
- **`cur_col_expr`** (`list` of `expresssion` objects) - symbolic
  expression for subset corresponding to current column.
- **`cur_col_n`** (`integer`) - column count for the current column
- **`cur_col_split`** (`character`) - split name that resulted in the
  current column
- **`cur_col_split_value`** (`character`) - split value (for
  `cur_col_split`) that resulted in the current column
  

# Designing Conditional Behavior in `afun`s

Recall that table contents are (typically) calculated by repeated
calling the analysis or content function for a given row facet - once
per individual column within the table structure.

## Conditional-On-Column Behavior in `afun`s

If we want our table to have different *types* of content in
differnt cells of the same row, we need an `afun` that

1. can determine where in column space it is calculating cells for, and
2. implements two or more behaviors which it selects between based on (1)

**Note** it is *mandatory* that the calls to our analysis or content
function result in the same number of rows within each column. This
can involve padding the results with blank cells in some columns.
