---
title: "Advanced rtables - Understanding and Interacting With `TableTree` Objects"
subtitle: Contributed by Johnson & Johnson Innovative Medicine
date: "2025-10-22"
author:
- Gabriel Becker
- Dan Hofstaedter
output:
    rmarkdown::html_document:
        theme: "spacelab"
        highlight: "kate"
        toc: true
        toc_float: true
        code_folding: show
vignette: >
  %\VignetteIndexEntry{Advanced rtables - Understanding and Interacting With TableTree Objects}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
      wrap: 72
  chunk_output_type: console
---

```{r, include = FALSE}
suggested_dependent_pkgs <- c("dplyr")
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = all(vapply(
    suggested_dependent_pkgs,
    requireNamespace,
    logical(1),
    quietly = TRUE
  ))
)
```

```{r, echo=FALSE}
knitr::opts_chunk$set(comment = "#")
```

# Coming Soon

  - [Table Structure and Pathing](./pathing.html) Table structure and
    describing locations within a table via pathing (existing
    vignette)
  - [Accessing Values Within A
    Table](./guided_advanced_tt_access.html) Retrieving values
    from a table
  - [Writing Custom Scoring Functions For
    Sorting](./guided_advanced_tt_score_funs.html) Writing custom
    scoring functions for use with `sort_at_path`
  - [Writing Custom Pruning
    Functions](./guided_advanced_tt_prune_funs.html) Writing custom
    pruning functions for use with `prune_table`
