---
title: "Introspecting tables"
author: "Gabriel Becker and Adrian Waddell"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introspecting tables}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r, echo=FALSE}
knitr::opts_chunk$set(comment = "#")
```

```{css, echo=FALSE}
.reveal .r code {
    white-space: pre;
}
```

The packages used for this vignette are `rtables` and `dplyr`:

```{r, message=FALSE}
library(rtables)
library(dplyr)
```

## Introduction

First, let's set up a simple table.

```{r}
lyt <- basic_table(show_colcounts = TRUE) %>%
  split_cols_by("ARMCD") %>%
  split_cols_by("STRATA2") %>%
  split_rows_by("STRATA1") %>%
  add_overall_col("All") %>%
  summarize_row_groups() %>%
  analyze("AGE", afun = max, format = "xx.x")

tbl <- build_table(lyt, ex_adsl)
tbl
```

## Getting Started

We can get basic table dimensions for the number of rows and columns with 
the following code:
```{r}
dim(tbl)
nrow(tbl)
ncol(tbl)
```

## Detailed Table Strucutres

The `table_structure()` function prints a summary of a table's row
structure at one of two levels of detail. By default, it summarizes
the structure at the subtable level.

```{r}
table_structure(tbl)
```

When the `detail` argument is set to `"row"`, however, it provides a
more detailed row-level summary, which acts as a useful alternative to
how we might normally use the `str()` function to interrogate compound
nested lists.

```{r}
table_structure(tbl, detail = "row")
```

The `make_row_df()` and `make_col_df()` functions create a data.frame
which has a variety of information about the table's structure. Most
useful for introspection purposes are the `label`, `name`,
`abs_rownumber`, `path` and `node_class` columns (the remainder of
information in the returned data.frame is used for pagination)

```{r}
make_row_df(tbl)[,c("label", "name", "abs_rownumber", "path", "node_class")]
```

There is also wrapper available for `make_row_df` to display only the row path
structure:
```{r}
row_paths(tbl)
```

By default `make_row_df()` summarizes only visible rows, but setting
`visible_only` to `FALSE` gives us a structural summary of the table,
including the full hierarchy of subtables, including those that aren't
represented directly by any visible rows:

```{r}
make_row_df(tbl, visible_only = FALSE)[, c("label", "name", "abs_rownumber", "path", "node_class")]
```

`make_col_df()` similarly accepts `visible_only`, though here the
meaning is slightly different, indicating whether only *leaf* columns
should be summarized (`TRUE`, the default) or whether higher level
groups of columns, analogous to subtables in row space, should be
summarized as well.

```{r}
make_col_df(tbl)[, c("label", "name", "abs_pos", "path", "leaf_indices")]
```

```{r}
make_col_df(tbl, visible_only = FALSE)[, c("label", "name", "abs_pos", "path", "leaf_indices")]
```

Similarly, there is a wrapper available displaying only the column structure:
```{r}
col_paths(tbl)
```

The `row_paths_summary()` and `col_paths_summary()` functions wrap the
respective `make_*_df` functions, printing the `name`, `node_class`
and `path` information (in the row case), or the `label` and `path`
information (in the column case), indented to illustrate table
structure:

```{r}
row_paths_summary(tbl)
```


```{r}
col_paths_summary(tbl)
```

## Applications
Knowing the structure of an `rtable` object is helpful for retrieving specific
values from the table. 
For examples see [`subsetting_tables`](https://insightsengineering.github.io/rtables/main/articles/subsetting_tables.html#path-based-cell-value-accessing) vignette.

Understanding the table structure is also important for post-processing tables
such as sorting and pruning. More details on this are covered in the [`sorting_pruning`](https://insightsengineering.github.io/rtables/main/articles/sorting_pruning.html)
vignette.


## Summary

In this vignette learned that:

* A number of utility functions are available for examining the underlying structure
of an `rtables` object.