---
title: "Constructing rtables Manually"
author: "Adrian Waddell"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Constructing rtables Manually}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---


```{r, echo=FALSE}
knitr::opts_chunk$set(comment = "#")
```

```{css, echo=FALSE}
.reveal .r code {
    white-space: pre;
}
```

## Overview

The main functions currently associated with `rtable`s are

Tables in `rtables` can be constructed via the layout or `rtabulate` tabulation frameworks or also manually. Currently manual table construction is the only way to define column spans. The main functions for manual table constructions are:

* `rtable`: collection of `rrow` objects, column header and default format
* `rrow`: collection of `rcell` objects and default format
* `rcell`: collection of data objects and cell format

## Simple Example


```{r, message=FALSE}
library(rtables)
```
```{r}
tbl <- rtable(
  header = c("Treatement\nN=100", "Comparison\nN=300"),
  format = "xx (xx.xx%)",
  rrow("A", c(104, .2), c(100, .4)),
  rrow("B", c(23, .4), c(43, .5)),
  rrow(),
  rrow("this is a very long section header"),
  rrow("estimate", rcell(55.23, "xx.xx", colspan = 2)),
  rrow("95% CI", indent = 1, rcell(c(44.8, 67.4), format = "(xx.x, xx.x)", colspan = 2))
)
```


Before we go into explaining the individual components used to create this table
we continue with the html conversion of the `rtable` object:


```{r}
as_html(tbl, width = "80%")
```

Next, the `[` operator lets you access the cell content.

```{r}
tbl[1, 1]
```

and to format that cell run `format_rcell(tbl[1,1])`=`r #format_rcell(tbl[1,1])`.

Note that `tbl[6, 1]` and `tbl[6, 2]` display both the same `rcell` because of
the `colspan`.



