---
title: "rtabulate"
author: "Adrian Waddell"
date: "3/21/2018"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, echo = FALSE}
library(rtables)
```

## Introduction

`rtabulate` provides a number of methods to compute `rtables`. The idea behind
`rtabulate` is to have a similar funtion to `tapply` that returns rtables
instead of lists. For example, for the iris data we might be interested in the
average `Sepal.Length` for each `Species` which can be answered with `tapply` as
follows:

```{r}
attach(iris)
tapply(X = Sepal.Length, INDEX = Species, FUN = mean)
```

and to get the same numbers with `rtabulate` we can run

```{r}
rtabulate(x = Sepal.Length, col_by = Species, FUN = mean)
```

Because rtables also provide formatting options we can pass that info to
`rtabulate`

```{r}
rtabulate(x = Sepal.Length, col_by = Species, FUN = mean, format = "xx.xx")
```

This vignette will provide an overview of `rtabulate` and show how different
types of tables can be derived with it.


## Tabulation Methods

The tabulation methods follow the split apply combine pattern. That is, for a
vector `x`, a factor `col_by`, and a function `FUN` the steps are:

1. split `x` by `col_by`
   ```{r, eval = FALSE}
   x_s <- split(x, col_by)
   ```
1. apply the function `FUN` to each element, e.g. `FUN(x_s[[1]])`, and  in the split data which is then
treated as an `rcell`

1. combine the reclls to an rrow and create an rtable




### Tabulating Numerical Data

### Tabulating Boolean Data

### Tabulating Factors

### Tabulation based on Data Frames

