---
title: "Introduction to rtables"
author: "Adrian Waddell"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

# Overview

`rtable` is useful to create tables where multiple values are stored in a cell 
with a particular format. An `rtable` object has access to the individual raw 
values and prints the data with formatting rules such as rounding and pasting.
`rtables` also allow for merging cells within a row.

The main functions currently associated with `rtable`s are

* `rtable`: collection of `rrow` objects, column header and default format
* `rrow`: collection of `rcell` objects and default format
* `rcell`: collection of data objects and cell format
* `as_html` and `toString`: output an `rtable` as ascii or html

# Simple Example


```{r, message=FALSE}
library(rtables)
```
```{r}
tbl <- rtable(
  col.names = c("Treatement\nN=100", "Comparison\nN=300"),
  format = "xx (xx.xx%)",
  rrow("A", c(104, .2), c(100, .4)),
  rrow("B", c(23, .4), c(43, .5)),
  rrow(),
  rrow("this is a very long section header"),
  rrow("estimate", rcell(55.23, "xx.xx", colspan = 2)),
  rrow("95% CI", indent = 1, rcell(c(44.8, 67.4), format = "(xx.x, xx.x)", colspan = 2))
)

tbl  
```


Before we go into explaining the individual components used to create this table
we continue with the html conversion of the `rtable` object:


```{r}
as_html(tbl, width = "80%")
```

Next, the `[` operator lets you access the cell content.

```{r}
tbl[1, 1]
```

and to format that cell run `format_rcell(tbl[1,1])`=`r format_rcell(tbl[1,1])`.

Note that `tbl[6, 1]` and `tbl[6, 2]` display both the same `rcell` because of
the colspan.

# Viewer

The `Viewer` functions takes an `rtable` object and displays it with
[bootstrap](https://getbootstrap.com/) styling in RStudio's Viewer pane.

```{r, eval = FALSE}
Viewer(tbl)
```

It is also possible to show two tables side by side

```{r, eval = FALSE}
Viewer(tbl, tbl)
```


# Formats

Cell formats are inherited from `rtable` to `rrow` to `rcell` if there is not a
more specific format argument (i.e. format arguments of `rcell` are more
specific than those of `rrow` which are more specific than those of `rtable`.

Two types of formats are supported:

* format label, e.g. `xx.xx (xx.xx%)`

* functions that take the element and output type as an argument and return a 
formatted string or html object **note, we will eventually deprecate this**

Formats labels are currently limited to the following predefined set:

```{r}
get_rcell_formats()
```

Here is an example of using functions for formatting cells:

```{r}
my_format <- function(x, output) {
   paste(x, collapse = "/")
}
tbl3 <- rtable(
  c("A", "B"),
  format = my_format,
  rrow("row1", c(1,2,3,4), letters[1:10])
)
tbl3
```



# Comparing two rtables

Because we have the original data for each cell accessible for the `rtable` data
structure it is possible to programmatically compare two tables:

```{r}
t1 <- rtable(col.name = c("A", "B"), format = "xx", rrow("row 1", 1, 2))
t2 <- rtable(col.name = c("A", "B", "C"), format = "xx", rrow("row 1", 1, 2, 3))

compare_rtables(object = t1, expected = t2) 
```



# Converting table objects

You can also convert table objects to rtables:

```{r}
as.rtable(table(iris$Species))
```

or 

```{r}
print(as.rtable(with(mtcars, table(cyl, hp))), gap = 1)
```
