---
title: "rtabulate"
author: "Adrian Waddell"
date: "3/21/2018"
output:
  pdf_document: default
  html_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, echo = FALSE}
library(rtables)
```

## Introduction

`rtabulate` provides a number of methods to compute `rtable` tables. The idea
behind `rtabulate` is to have a similar function to `tapply` which returns
rtables instead of lists. For example, for the iris data we might be interested
in the average `Sepal.Length` for each `Species` which can be answered with
`tapply` as follows:

```{r}
attach(iris)
tapply(X = Sepal.Length, INDEX = Species, FUN = mean)
```

and to get the same numbers with `rtabulate` we can run

```{r}
rtabulate(x = Sepal.Length, col_by = Species, FUN = mean)
```

Because rtables also provide formatting options we can pass that info to
`rtabulate`

```{r}
rtabulate(x = Sepal.Length, col_by = Species, FUN = mean, format = "xx.xx")
```

This vignette will provide an overview of `rtabulate` and show how different
types of tables can be derived with it.


## Tabulation Methods

The tabulation methods follow the split apply combine pattern. That is, for a
vector `x`, a factor `col_by`, and a function `FUN` the steps are:

1. split `x` by `col_by`
```{r, eval = FALSE}
x_s <- split(x, col_by)
```
   
1. apply the function `FUN` to each element, e.g. `FUN(x_s[[1]])`,
`FUN(x_s[[2]])`, etc...,  which either should return an `rcell` or a data
structure that gets wrapped into an `rcell`

1. combine the rcells to an rrow and create an rtable with a header according to
the levels of `col_by`


If `x` should not be split then the `no_by` function can be used to specify a
column name (as the returned object is still an rtable with one column):

```{r}
rtabulate(Sepal.Length, no_by("All Data"), sum)
```


### Cell Data Based Tabulation

#### Tabulating Numerical Data

`rtabulate.numeric` returns an `rtable` with one row and columns according to
the levels of `col_by`.

```{r}
rtabulate(x = Sepal.Length, col_by = Species, FUN = median)
```

#### Tabulating Boolean Data

Like `rtabulate.numeric`, `rtabulate.logical` returns an rtable with one row and
columns according to the levels of `col_by`.


```{r}
is_SL_below_avg <- Sepal.Length < mean(Sepal.Length)

rtabulate(x = is_SL_below_avg, col_by = Species, row.name = "count (percentage)")
```

By default `FUN=function(x) sum(x) * c(1, 1/length(x))`.

#### Tabulating Factors

`rtabulate.factor` returns an `rtable` with one row per factor level of `x`.

```{r}
rtabulate(x = esoph$agegp, col_by = esoph$alcgp)
```

#### Tabulation based on Data Frames

`rtabulate.data.frame` has a `data.frame` as `x` argument and the `col_by_var`
and `row_by_var` need to specify which factor variables with `x` should be used
to split the data for the table rows and columns. By default the argument to
`FUN` is then a subset of `x` that is associated to a particular table cell.
Let's look at the `CO2` dataset:

```{r}
head(CO2)
```

Say we want to calculate the total `uptake` for each `Type` and `Treatment`

```{r}
rtabulate(x = CO2, row_by_var = "Type", col_by_var = "Treatment",
          FUN = function(xi) sum(xi$uptake))
```


### Cell Data, Row Data, Column Data based Tabulation

`rtabulate.numeric` and `rtabulate.logical` have the argument `row_data_arg`
which is by default set to `FALSE`. If set to `TRUE` then `FUN` receives a
second argument with a copy of the `x` argument (as the row-associated data is
the whole data for a `numeric` and `logical` vector).


```{r}
rtabulate(
  Sepal.Length, Species,
  FUN = function(x_cell, x_row) c(length(x_cell), length(x_row)),
  row_data_arg = TRUE,
  format = "xx / xx", 
  row.name = "length of cell and row data"
)
```


Next, for `rtabulate.factor` and `rtabulate.data.frame` there is an
`row_col_data_args` which when set to `TRUE` then the `FUN` function receives
three argument, the data that is associated to a table cell, row, column,
respectively.

```{r}
mtcars2 <- mtcars
mtcars2$gear <- factor(mtcars2$gear)
mtcars2$carb <- factor(mtcars2$carb)

rtabulate(
  x = mtcars2, row_by_var = "carb", col_by_var = "gear",
  FUN = function(x_cell, x_row, x_col) c(nrow(x_cell), nrow(x_row), nrow(x_col)),
  row_col_data_args = TRUE,
  format = NULL
)
```

Note that `format=NULL` is equivalent to `paste(x, collapse = ", ")` on the cell
data structure.

