---
title: "rtabulate"
author: "Adrian Waddell"
date: "3/21/2018"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rtabulate}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, echo = FALSE}
library(rtables)
```

## Introduction

`rtabulate` provides a number of methods to compute `rtable` tables. The idea
behind `rtabulate` is to have a similar function to `tapply` which returns
rtables instead of lists or vectors. For example, for the iris data we might be
interested in the average `Sepal.Length` for each `Species` which can be
answered with `tapply` as follows:

```{r}
attach(iris)
tapply(X = Sepal.Length, INDEX = Species, FUN = mean)
```

and to get the same numbers with `rtabulate` we can run

```{r}
rtabulate(x = Sepal.Length, col_by = Species, FUN = mean)
```

Because rtables also provide formatting options we can pass that info to
`rtabulate`

```{r}
rtabulate(x = Sepal.Length, col_by = Species, FUN = mean, format = "xx.xx")
```

This vignette will provide an overview of `rtabulate` and show how different
types of tables can be derived with it.


## Tabulation Methods

The tabulation methods follow the split apply combine pattern. That is, for a
vector `x`, a factor `col_by`, and a function `FUN` the steps are:

1. split `x` by `col_by`
```{r, eval = FALSE}
x_s <- split(x, col_by)
```
   
1. apply the function `FUN` to each element, e.g. `FUN(x_s[[1]])`,
`FUN(x_s[[2]])`, etc...,  which either should return an `rcell` or a data
structure that gets wrapped into an `rcell`

1. combine the rcells to an rrow and create an rtable with a header according to
the levels of `col_by`


If `x` should not be split, i.e. all should be taken, then the `by_all` function can be used to specify a
column name (as the returned object is still an rtable with one column):

```{r}
rtabulate(Sepal.Length, by_all("All Data"), sum)
```


### Cell Data Based Tabulation

#### Tabulating Numerical Data

`rtabulate.numeric` returns an `rtable` with one row and columns according to
the levels of `col_by`.

```{r}
rtabulate(x = Sepal.Length, col_by = Species, FUN = median)
```

#### Tabulating Boolean Data

Like `rtabulate.numeric`, `rtabulate.logical` returns an rtable with one row and
columns according to the levels of `col_by`.


```{r}
is_SL_below_avg <- Sepal.Length < mean(Sepal.Length)

rtabulate(x = is_SL_below_avg, col_by = Species, row.name = "count (percentage)")
```

By default `FUN=function(x) sum(x) * c(1, 1/length(x))`.

#### Tabulating Factors

`rtabulate.factor` returns an `rtable` with one row per factor level of `x`.

```{r}
rtabulate(x = esoph$agegp, col_by = esoph$alcgp)
```

#### Tabulation based on Data Frames

`rtabulate.data.frame` has a `data.frame` as `x` argument and the `col_by` and
`row_by` need to specify how to split the data for the table rows and columns.
The argument to `FUN` is then a subset of `x` that is associated with a
particular table cell. Let's look at the `CO2` dataset:

```{r}
head(CO2)
```

Say we want to calculate the total `uptake` for each `Type` and `Treatment`

```{r}
rtabulate(x = CO2, row_by = CO2$Type, col_by = CO2$Treatment,
          FUN = function(xi) sum(xi$uptake))
```


### Column Wise Arguments

The arguments passed to the elipsis `...` in `rtabulate` are optional arguments
to `FUN`. This is the same behaviour as for the arguments passed to `...` in
`lapply` and `tapply`. This is useful, for example, when passing the `na.rm`
argument to `mean`

```{r}
x <- c(1, 2, 3, NA, 3, 2)
cb <- factor(c("A", "B", "A", "B", "B", "A"))

rtabulate(x, cb, mean, na.rm = TRUE)
```


Often, howewer, it is also useful to iterate over multiple collections as for
example is the case with the `...` in `mapply` and `Map`. That is, given a
collection `x` and `y` with each `p` elements, then we would like to evaluate
`FUN(x[i], y[i])` for `i` in `1` to `p`. In `rtabulate` this can be achieved
with the collections wrapped in a named list and passed to the `col_wise_args`.
For example:

```{r}
x <- c(1, 2, 3, NA, 3, 2)
cb <- factor(c("A", "A", "A", "B", "B", "A"), levels = c("A", "B"))

tbl <- rtabulate(x, cb, FUN = function(xi, na.rm, N) {
  list(mean(xi, na.rm=na.rm), N)
}, na.rm = TRUE, col_wise_args = list(N = table(cb)), row.name = "mean, N")

tbl
```

Note that `format=NULL` is equivalent to `paste(x, collapse = ", ")` on the cell
data structure.

In order to add a `(N=xx)` type of row to the header we can add `header_add_N` function:

```{r}
header_add_N(tbl, table(cb))
```
