---
title: "Subsetting and Manipulating Table Contents"
author: "Gabriel Becker and Adrian Waddell"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Subsetting and Manipulating Table Contents}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---


```{r, echo=FALSE}
knitr::opts_chunk$set(comment = "#")
```

```{css, echo=FALSE}
.reveal .r code {
    white-space: pre;
}
```

## Introduction

`TableTree` objects are based on a tree data structure as the name indicates. The package is written such that the user
does not need to walk trees for many basic table manipulations. Walking trees will still be necessary for certain
manipulation and will be the subject of a different vignette.

In this vignette we show some methods to subset tables and to extract cell values.

We will use the following table for illustrative purposes:

```{r}
library(rtables)
library(dplyr)

tbl <- basic_table() %>%
  split_cols_by("ARM") %>%
  split_rows_by("SEX", split_fun = drop_split_levels) %>%
  analyze(c("AGE", "STRATA1")) %>%
  build_table(ex_adsl %>% filter(SEX %in% c("M", "F")))

tbl
```

## Traditional Subsetting and modification with `[`

The `[` and `[<-` acessor functions operate largerly the same as their `data.frame` cousins:

 - `[` and `[<-` both treat Tables as rectangular objects (rather than trees)
   - In particular this means label rows are treated as rows with empty cell values, rather than rows without cells
 - `[` accepts both column and row absolute position, and missing arguments mean "all indexes in that dimension"
   - multiple values can be specified in both row and column position
   - negative numeric positions are supported, though like `[.data.frame` they cannot be mixed with positive ones
 - `[` always returns the same class as the object being subset unless `drop = TRUE`
 - `[ , drop = TRUE` returns the raw (possibly multi-element) value associated with the cell.
 

**Known Differences from `[.data.frame`**

- absolute position cannot currently be used to reorder columns or rows
  - Note in general the result of such an ordering is unlikely to be structurally valid
- `character` indices are treated as paths, not vectors of names in both `[` and `[<-`

The `[` accessor function always returns an `TableTree` object if `drop=TRUE` is not set. The first argument are the row indices and the second argument the column indices. Alternatively logical subsetting can be used. The indices are based on visible rows and not on the tree structure. So:

```{r}
tbl[1, 1]
```

is a table with an empty cell  because the first row is a label row. We need to access a cell with actual cell data:

```{r}
tbl[3, 1]
```

To retrieve the value,  we use `drop = TRUE`:

```{r}
tbl[3, 1, drop = TRUE]
```


One can access multiple rows and columns:

```{r}
tbl[1:3, 1:2]
```


Note that we do not repeat label rows for descending children, e.g.

```{r}
tbl[2:4, ]
```

does not show that the first row is derived from `AGE`. In order to repeat content/label information one should use the pagination feature. Please read the related vignette.

Character indices are interpreted as paths (see below), NOT elements to be matched against `names(tbl)`:

```{r}
tbl[,c("ARM", "A: Drug X")]
```



## Path Based Cell Value Accessing:
Tables can be subset or modified in a structurally aware manner via *pathing*. 

Paths define semantically meaningful positions within a constructed table that correspond to the logic of the layout
used to create it.

A path is an ordered set of split names, the names of subgroups generated by the split, and the `@content` directive, which steps itno a position's content (or row group summary) table.

We can see the row and column paths of an existing table via the `row_paths`, `col_paths`, `row_paths_summary`, and `col_paths_summary`, functions, or as a portion of the more general `make_rows_df` function output.


```{r}


tbl2 <- basic_table() %>%
  split_cols_by("ARM") %>%
    split_cols_by("SEX", split_fun = drop_split_levels) %>%
    split_rows_by("RACE", split_fun = drop_split_levels) %>%
    summarize_row_groups() %>%
    analyze(c("AGE", "STRATA1")) %>%
    build_table(ex_adsl %>% filter(SEX %in% c("M", "F") &
                                   RACE %in% (levels(RACE)[1:3])))

tbl2
```

So the column paths are as follows:

```{r}
col_paths_summary(tbl2)
```

and the row paths are as follows:

```{r}
row_paths_summary(tbl2)
```

To get a semantically meaningful subset of our table, then, we can use `[` (or `tt_at_path` which underlies it)

```{r}

tbl2[c("RACE", "ASIAN"), c("ARM", "C: Combination")]
```

We cal also retrieve individual cell-values via the `value_at` convenience function, which takes a pair of row and column paths which resolve together to an individual cell, e.g. average age for asian femal patients in arm A:

```{r}
value_at(tbl2, c("RACE", "ASIAN", "AGE",  "Mean"), c("ARM", "A: Drug X", "SEX", "F"))
```

You can also request information from non-cell specific paths with the `cell_values` function:

```{r}
cell_values(tbl2, c("RACE", "ASIAN", "AGE", "Mean"), c("ARM", "A: Drug X"))
```

Note the return value of `cell_values` is always a list even if you specify a path to a cell:

```{r}
cell_values(tbl2, c("RACE", "ASIAN", "AGE",  "Mean"), c("ARM", "A: Drug X", "SEX", "F"))
```

## Rows and cell values alignments

It is possible to align the content of both rows (`in_rows`) and cell values (`rcell`) by assigning `"left"`, `"center"` (default), and `"right"` to `.aligns` and `align` arguments, respectively.

Please consider using `?in_rows` and `?rcell` for further clarifications on the two arguments. Also, to see all available alignment options, run the following:
```{r}
rtables_aligns()
```
Example that uses different `align` arguments for each `rcell`:
```{r}
lyt <- basic_table() %>%
    analyze("AGE", function(x) {
        in_rows(left = rcell("l", align = "left"),
                right = rcell("r", align = "right"),
                center = rcell("c", align = "center"))
    })

build_table(lyt, DM)
```

Example that uses `.aligns` argument to set all the rows (`in_rows`) to a single alignment format:
```{r}
lyt2 <- basic_table() %>%
    analyze("AGE", function(x) {
        in_rows(left = rcell("l"),
                right = rcell("r"),
                center = rcell("c"), .aligns = "right")
    })

build_table(lyt2, DM)
```
