---
title: "TableTree Design"
author: "Gabriel Becker"
date: "November 2019"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{rtabulate}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, echo = FALSE}
library(rtables)
```

# General Definitions

These definitions will be used throughout this document. Other
sections may contain further definitions specific to their topic

- Table - An collection of Rows which share a common column structure,
  with a unique ordering of rows defined on it. These can have 2 forms
  - Flat table - ordered list of rows which define a standard
    rectangular dataset/table
  - Table tree - A nested hierarchical structure in which nesting
    represents relationships between rows. 
- Row - a single row in the table, can be the following types
  - Label row - a row with no data there to display a variable or
    other label for a position in the nesting structure
  - Content row - a row within the content table at a position within
    the tree, contains summary or aggregate data for that nesting
    level
  - Data row - a row containing the result(s) of the tabulation (e.g.,
    mean of AGE) for a subset of the data defined by position in the
    tree structure
- Tabulation functions - A function which is applied to appropriate
  subsets of data in order to generate row contents
  - Analysis function - A tabulation function used to generate the
    values for one or more _Data_ rows
  - Summary Function - A tabulation function used to generate values
    for a _Content_ row.
- Comparison Tabulation - Tabulation which compares the data subsets
  corresponding to 2 (or more??? not currently) columns in the table.
- Comparison function - A function which accepts one or two objects
  calculated from the data subsets being compared
  - Direct Comparison Function - a comparison function which takes two
    values, one calculated from each subset under comparison, and
    returns values which should populate the data cell
  - Table-based Comparison Function - a comparison function which
    accepts a 2d (2xk) contingency table representing the subset
    (baseline vs comparison column) by the value of the variable being
    analyzed.


# Layouts

Layouts are declared *pre-data* as the structure the table and
associated aggregation will have once it is applied to data.

This allows us to re-use layouts for different datasets (that have
the same relevant structure) and is a core design choice.

## Building up Layouts

Layouts define both the column- and row-based nesting structure and
are built up by piping together a declarative "grammar of table
structure".

The core design here is that adding more structure will add nesting by
default, unless adding structure where nesting does not make sense
(e.g., "total" or "all" column) or the user indicates otherwise.

So if

`NULL %>% add_colby_varlevels("ARM")`

results in a table like 

```
       A    B
Value  5    8
```

Then

`NULL %>% add_colby_varlevels("ARM") %>% add_colby_varlevels("SEX")`

would give us something like

```
          A      B
        M  F    M  F
Value   1  5    3  3
```

This is intended to give users/developers a very convenient, intuitive
syntax for building up complex table structures using basic components.

We can always bypass the default if nesting is not what is desired, for
example

`NULL %>% add_colby_varlevels("ARM") %>% add_colby_varlevels("SEX", newtoplev = TRUE)`

Would give us a column structure like

```
       A   B   M   F
Value  5   6   4   7
```

If the user did ever want that (this type of multiple parallel top-level
sub-roots structure will be much more common in row space than column
space by my estimation).


# The Object Model
## Core Pieces - Splits

Table structure is modeled via declaring "splits", formally
modeled by the `Split` S4 virtual class and its non-virtual
subclasses.

At its core, a Split declares a piece of how the final set of
rows or columns of a table will be generated.

`Split` objects are declarative; we don't require data to
construct them. This means that they cannot know their set
of children when built and thus are pieces of blueprints
for a table structure, rather than a structure themselves.

`Split` objects can be placed together, predata to form
`SplitVector` objects (just a list of `Split`s which define
a nested (sub) structure in row or column space.

### Non-comparison Split Types
- `VarLevelSplit` - Will generate children for each subset of a
  categorical variable
- `AllSplit` - Will generate one child containing the full set of data
  (at the current level of nesting)
- `RootSplit` - An AllSplit that is explicitly the root of the full
  row/column structure for the table
- `MultiVarSplit` - Will generate a child for each specified variable
- `VarStaticCutSplit` - Will generate children based on 'cutting' the
  values of a specified variable based on pre-specified points
- `VarDynCutSplit` - Will generate children based on 'cutting' the
  values of a specified variable via a function at aggregation time
- `AnalyzeVarSplit` - Declares that splitting is done and aggregation
  should occur at this stage by applying the specified function to the
  values of the specified variable
- `SubsetSplit` - Will generate one child based on inclusion in a
  subset (indicated by values of a logical variable). Additional
  child/children will be added for at least one of: Non-inclusion, all
  observations.

### Comparison Splits

Comparison splits declare that two intermediate aggregation values
should be generated and then computed on together to generate the
value of a cell.

E.g,

`
NULL %>% add_colby_blinecomp(var = "visit",
     	 	             baseline = "BASE",
			     comparison = '-',
			     rawvalcol = TRUE) %>%
      add_analyzed_var("BP", "Blood pressure)
`

Would give us something like

```
      	        BASE    VISIT1     (VISIT1 - BASE)    VISIT2     (VISIT2 - BASE)
Blood pressure  80      83          3                 98         15
```

*NB* These are subtractions of the aggregate values, NOT aggregates of
     the patient level differences. This honestly may not be very
     useful...

XXX I have an email in to Adrian about whether this is even something
    we want. If we want aggregation of patient level comparisons and
    the comparison is already calculated this is way easier
      
      
      AW: why don't you just ask for a function with two arguments
       `xb`, `xc` so that it is left to the user to define what to 
       calculate?
       The baseline changes are usually calculated before tabulations\
       but test based comparisons are not.
       

## Populated Table Structure

A table tree is a nested structure of `TableTree` objects, each of
which has a `content` `ElementaryTable` and a set of children (which
are themselves either `TableTree` or `TableRow` objects). An
ElemntaryTable is a table structure with no content slot and whose
children must be rows (not nested tables).

Position (on either of the table's axes) is tracked by a `TreePos` (or
subclass) object. A `TreePos` contains a list of Splits and a parallel
list of `SplitValue` objects which define the child "selected" at each
split to get to the current position. Each subtree and row within the
full tree is aware of its position (in row space).

### InstantiatedColumnInfo

A populated table (and all of its subtables) carry around an
`InstantiatedColumnInfo` object, which contains all of the various
representations of the column structure necessary for the machinery to
work

This includes

- A `LayoutColTree` object which contains the tree/graph structure of
  the hierarchical columns
- A list of subsetting expressions for each "leaf column", ie the
  columns ultimately representing the length of the row data vectors
  (leaves of the tree representation)
- a list of extra arguments associated with each leaf column.
- a vector of counts for the leaf columns
- Whether the counts should be displayed, and if so what format should
  be used.

We can look at a table's column info via the `col_info` accessor (at
time of writing there's no nice show method, but there likely will be
someday).

NB: This has grown organically over time, and honestly it is an open
question how much we actually need the tree structure representation
for once a table has been created from layout + data.



## Pre-Data


The table structure is declared prior to data being present (and thus
can be re-used).

This means that in the pre-data stage we cannot actually build the
nesting and children yet.

### Pre-data Layouts

A `PreDataTableLayouts` object is built by a layout pipe chain
(discussed below) and contains a `PreDataColLayout` and a
`PreDataRowLayout`. `PreDataColLayout` and `PreDataRowLayout` objects
are both lists of one or more `SplitVector` objects declaring nesting
substructures. Each also contains a `RootSplit` object though that is
not likely to affect anything in most cases.

# Declaring layout

Table structure is declared (pre-data) via a set of sequential
`add_colby_*`, `add_rowby_*`, `add_analyzed_var` and `add_summary_*`
calls within a layout pipeline (which currently starts with NULL as an
artifact).

At each stage in the pipeline, the next action occurs *at the current
level of nesting* unless otherwise indicated by the user or required
definitionally.

For example

```
layout = NULL %>% add_colby_varlevels("ARM", "Arm") %>%
       add_colby_varlevels("SEX", "Gender") %>%
       add_summary_count(lbl = "Overall (N)") %>%
       add_rowby_varlevels("RACE", "Ethnicity") %>%
       add_summary_count("RACE", lblfmt = "%s (n)") %>%
       add_analyzed_var("AGE", "Age", afun = mean, fmt = "xx.xx")
```

Will generate a layout where each arm of the trial has a column for
each value of the SEX variable present (in the overall dataset, not in
that arm). For row structure, it will have a top level `content` row
with overall count, then for each level of RACE, it will have a sub
count, then an `analysis` of AGE (the mean) within that ethnicity.

``` 
> layout A Pre-data Table Layout

Column-Split Structure:
ARM (lvls) -> SEX (lvls) 

Row-Split Structure:
RACE (lvls) -> AGE (** analyzed var **) 
```

Note this layout is then usable with any dataset that has the `ARM`,
`SEX`, `RACE`, and `AGE` variables present.


## Analyzed Variables


Adding /analyzed variables/ to our table layout defines the primary
tabulation to be performed. We do this by adding calls to
`add_analyzed_var` and/or `add_analyzed_colvar` into our layout
pipeline. As with adding further splitting, the tabulation will occur
at the current/next level of nesting by default, so

```
layout <- NULL %>%  %>% add_colby_varlevels("ARM", "arm") %>%
     add_colby_varlevels("SEX", "gender") %>%
     add_analyzed_var("AGE", summary)

``` 

Will tabulate values of `AGE` after grouping/splitting by `ARM` and
then `SEX` by calling `summary()`.

## Tabulation functions

Tabulation functions should return a list or vector with one element
per row to be created. Row labels are taken from the names of the
list, so

```
tworows = function(x,...) {
		list(mean = mean(x),
	             sd = sd(x))
}
```

Will create _two_ rows, one for mean and one for sd, at each
tabulation position.

In order to have multi-valued cells, the tabulation function should
return a list containing vectors for any multi-valued rows. So to
create the standard mean(sd) syle row, instead of the above we would
do

```
onerow = function(x, ...) {
       list("mean (sd)" = c(mean = mean(x), sd = sd(x)))
}
```

### Tabulation Function Signatures

Tabulation functions, both analysis and summary, must take the data
to be analyzed as their first argument. This will either be a subset
data.frame (in the case of summary functions generating content rows)
or vector data (in the case of most analysis functions generating data
rows), depending on context.

Additionally, tabulation functions can optionally take 2 extra named
arguments: `.N_col` and `.N_total`. If these arguments are present
within the formals of the function, the number of observations
*represented overall in the current column*, and the number of
observations *total in the dataset*, will be passed to them,
respectively.

There is also currently support for tabulation functions being passed
column-specific `extra` arguments, which I will document here once it
is more concrete and actually used in the test script.

## Variable columns

In some cases, the variable to be ultimately analyzed is most
naturally defined on a column, not a row basis.

When we need columns to reflect _different variables entirely_, rather
than different levels of a single variable, we use
`add_colby_multivar`, which places a `MultiVarSplit` within the column
structure.

By default, this simply makes the subset associated with each column
the set of observations which are non-NA for that column's
variable. In order to maake the cell contents within the column
reflect tabulation/analysis of those variables, we need to use
`add_analyzed_colvars` instead of `add_analyzed_var` when specifying
the tabulation, like so.

```
layout = NULL %>% add_colby_varlevels("ARM", "Arm") %>%
       add_colby_multivar(c("value", "pctdiff")) %>%
       add_rowby_varlevels("RACE", "ethnicity") %>%
       add_analyzed_colvars(afun = mean)
```

The `add_analyzed_colvars` here indicates that the variables whose data
are ultimately processed by `afun` are specified at the highest-depth
level of nesting in the column structure, rather than at the row
level.

Each row can /either/ represent a tabulation of a single, row
associated variable across multiple column-associated data subsets, OR
represent the analysis/tabulation of multiple column associated
variables, but not both.

NOTE: it would be unusual, but is allowed, to mix rows that have each
of the above modalities in the same table. It is important to note,
however, that currently all rows in the same table MUST have the same
formal column structure, which would severely limit thee utility of
doing this in practice.


We may want to change `add_analyzed_var` to `add_analyzed_rowvar` if
we go this route to ensure clarity

Open question: does it ever make sense to have columns that specify a
variable AND analysis variables specified at the row level? I think
these would always be in conflict but need to think a bit harder
first.


## Column Counts

Column counts can be displayed a couple of ways, one is as a content
summary row on the root split, but the more common one is as part of
the header material. We declare this when building up the layout with
the`add_colcounts()` function. The default format is `"(N=xx)"` so
generally an empty call within the pipe workflow is sufficient.

```
layout = NULL %>% add_colby_varlevels("ARM", "Arm") %>%
       add_colby_multivar(c("value", "pctdiff")) %>%
       add_colcounts() %>%
       add_rowby_varlevels("RACE", "ethnicity") %>%
       add_analyzed_colvars(afun = mean)
```

The call can appear anywhere in the pipeline, but is perhaps easiest
to understand at the end of the column structure declaration part of
the pipeline, as above.

We can also change column count rendering behavior (they are always
calculated and available) after creating either a layout or a
TableTree via the `disp_ccounts<-` getter, like so

```
disp_ccounts(layout) = FALSE
tab = build_table(layout, dat)
disp_ccounts(tab) = TRUE
```



## Compound splits

The current design calls for layouts (in both the row and column
direction) to be lists of split vectors. Max brought up the
possibility of wanting to add splits at the current level of nesting,
instead of either descending or jumping back up to the top. I don't
want to complicate the pre-data layout structure by making it a full
tree, though.

Current plan is to support this with "compound splits" which are just
splits that hold 2 or more splits that should all be applied at the
given level of nesting.

NOTE: some possible issues with labeling in display.

One question is how to declare/add compound splits

### Option 1

```
NULL %>% add_colby_varlevels("ARM", "arm") %>%
     add_colby_compound(NULL %>% add_colby_all() %>%
     add_colby_varlevels("SEX", "sex")) %>%
     add_analyzed_var("AGE", summary)
```

Pros: no new syntax or grammar elements

Cons: I really hate the whole nesting pipes thing.
      A lot of useless work being done and undone internally here, the
      inner pipe would build up a PreDataLayouts object only to
      extract the split vector and ignore the rest.

### Option 2

```
NULL %>% add_colby_varlevels("ARM", "arm") %>%
     add_colby_compound(AllSplit(), VarLevelSplit("SEX", "gender")) %>%
     add_analyzed_var("AGE", summary)
```

Pros: no nested pipes. 
      Clear what it does. 

Cons: 
Not using the same declarative style grammer as everything
else. 'Manual' construction of individual split objects via
constructors

### Option 3

replace `newtoplevel` argument with `nestlevel` which can take:

- `"next"` (the default), 
- `"current"` (create or add to compound split at current nesting level), or
- `"top"` ( new top level)


```
NULL %>% add_colby_varlevels("ARM", "arm") %>%
     add_colby_all() %>% 
     add_colby_varlevels("SEX", "gender", nestlevel = "current") %>%
     add_analyzed_var("AGE", summary)
```

Pros: same ux/"formfeel" as the before
      users don't need to know about the concept of compound split objects at all

Cons: adds complexity to parameter space
      non-logical argument


### Different variables summarized per column

The above `add_analyzed_var` method for declaring tabulation does so
at the row level: each data row represents a (partial) tabulation of
single variable within the context of the different
columns. Sometimes, however, we want different columns to display
values/summaries of _different variables_. We do this by declaring the
columns via `add_colby_multivar` and then adding tabulation via
`add_analyzed_colvar`.

`add_analyzed_colvar` declares that a data cell's value is a
tabulation of the variable associated with the _column_ it is in,
grouped according to the row and column splitting, rather than
associating a variable with the row. Note that unlike
`add_analyzed_var` we do not specify a variable to be analyzed here,
only a tabulation function, as the column variable will have already
been declared via a call to `add_colby_multivar`. tabulation should be
done with respect to the variable/subset associated with the column,
rather than



# Comparison-based tabulation

Comparison-based values are where the tabulated value of a cell
depends on data from two (or more?) columns. For example, the
difference in mean measurements between the current column and a
declared baseline column.

Like single column tabulations, these comparison calculations are
declared without our layout and then performed automatically by the
tabulation machinery. In fact, currently they are implemented as a
special case of non-comparison tabulation.

## Declaring a column-split with a Baseline level

Which value should be considered the baseline when performing
automatic comparisons is done in the column split. Instead of doing
`add_colby_varlevels` we simply do `add_colby_varwbline` (the last
portion means "variable with baseline"), like so

```
layout = NULL %>% add_colby_varwbline("ARM", "Arm A", lbl = "Arm") %>%
       add_rowby_varlevels("RACE", "ethnicity") %>%
       add_analyzed_var("AGE", afun = mean)
```

Note that the above layout does not actually perform any comparisons,
because `add_analyzed_var` declares non-comparing tabulation.

For non-comparing tabulations `add_colby_varwbline` and the
corresponding `VarLevWBaselineSplit` split object which is added to
the layout, behave identically to `add_colby_varlevels` and its
corresponding `VarLevelSplit`.

NOTE: In light of the above, we can probably just always have the
deepest column-split have a baseline of the first level so people don't
even need to declare it. This is NOT what happens now, though, and
some care is needed because we only want comparisons at the deepest
level of nesting, I think. TODO(?)

## Declaring comparison-based tabulation

With a baseline value defined on our column split we can declare
comparison tabulations with our layout.

### Declaring direct comparisons

We declare direct comparisons with `add_analyzed_blcomp`, which take
`compfun` (defaulting to simple subtraction) in addition to `afun` so
to declare we want differences in means in our table we would do

```
layout = NULL %>% add_colby_varwbline("ARM", "Arm A", lbl = "Arm") %>%
       add_rowby_varlevels("RACE", "ethnicity") %>%
       add_analyzed_blinecomp("AGE", afun = mean)
```

Everything is then done by the rtables machinery internally and we get
a row with no value for Arm A, and the differences in mean between
Arm A and each of Arms B and C.

### Declaring Table-based comparisons

We provide a convenience function which declares that the comparisons
are table-based. In this case, calculating the contingency table is
done automatically and only the comparison function is specified by
the user, like so

```
layout = NULL %>% add_colby_varwbline("ARM", "Arm A", lbl = "Arm") %>%
       add_analyzed_blinecomp("rsp", 
       		compfun = function(tab) list("CI" = prop.test(tab)$conf.int*100))
```

See `tt_rsp` in `R/tt_rsp.R` for an example of table-based comparison
used in practice.

		     
# Manual Table Construction and manipulation

The primary aim of the class hierarchy laid out above is to declare,
and then automate the execution of, tabulations of raw data.

There are times, however, when a table with tabulated results may need
to be modified manually by results, or when a table must be
constructed manually without the use of a layout and the tabulation
machinery.

## Manual Construction

### Definitions

- Manual Split - a 'split' which defines the names of its children
  directly, and which does not meaningfully reflect a set of subsets
  of raw data. These should never be used in tabulation, but if they
  are, each child represents the entire subset of the data that
  "flowed into" the split.

### Column information/structure

`TableTree` objects, even manually constructed ones, have a formalized
column structure, distinct from the approach taken by the S3 `rtable`
object based approach.

Typically column structure in a `TableTree` reflects aspects of the
data and tabulation performed by the tabulation machinery, but the
structure can be declared manually in a data-independent way when
required.

We provide the `manual_cols` function which takes any number of
vectors of values, and uses those to construct a nested column
structure, with the column levels that come first in the call
containing those come later. For example, to get the familiar SEX
nested within ARM structure we have been using at places here, we
would do

```
manual_cols(ARM = c("ARM 1", "ARM 2"), SEX= c("M", "F"))
```

This constructs a full `InstantiatedColumnInfo` object suitable for
passing to the `TableTree` constructor.


### Constructing Elementary Tables

Constructing ElementaryTable objects manually is done like so

```
ElementaryTable(kids = list(TableRow(list(1, 2, 3, 4), lab = "Row 1"), 
                                TableRow(list(5, 6, 7, 8),
                                         lab = "Row 2")),
	        cinfo = manual_cols(ARM = c("ARM 1", "ARM 2"), SEX= c("M", "F")))
```

### Constructing Non-Elementary Tables

In general Tables in `rtables` are not Elementary, rather they are
`TableTree` objects. This means that there can be a nested row
structure (can have `TableTree` objects as children) and each
TableTree object has a content (elementary) table.

If nesting or content is required, we simply create them manually
using the TableTree, ElementaryTree, and TableRow constructors as
desired:

```
TableTree(cont = ElementaryTable(list(TableRow(list(20, 20, 15, 20)))),
          kids = list(subtab1 = TableTree(cont = ElementaryTable(list(TableRow(list(10, 10, 20 , 10)))),
	       	 	      		  kids = list(TableRow(list(5.5, 2.2, 5.5, 10.10)))),
                      subtab2 = ElementaryTable(list(TableRow(list(9, 90, 900, 9000))))),
          cinfo = manual_cols(ARM = c("ARM 1", "ARM 2"), SEX= c("M", "F")))

``` 

If the TableTree constructor is called but no nesting structure or
content rows/table are specified, an `ElementaryTable` object will be
constructed.


The `TableRow` constructor will directly create a `TableRow`
object. We simply pass it the values the row should have.


Typically we do not set column info when constructing a `TableRow`
because when added to a table without it, the rows inherit the table's
column structure.

### Constructing ElementaryTables


# Pagination

A first pass of the pagination algorithm is working and can be seen in
`tt_paginate.R`

Pagination is currently defined/modeled as a _rendering_ activity,
meaning it is /not/ formally modeled in the TableTree object
model. Rather, it is done transiently to a TableTree object in the
process of also doing other preparation for it to be printed. There
are some downsides to this but it is both much simpler than formally
modeling pagination and, I think, conceptually correct. I don't think
we want the TableTree objects themselves to know about pagination
because other output backends (e.g., html) will either not have
pagination at all or will handle it very differently.

## Definitions

- Header lines - The multi-row definition of the column layout and the
  dividing "row" displayed between the the columns and the table rows
- Repeat Row - A label or content row which must be repeated after
  pagination to correctly display the position/context of the first
  row in the new page.
- Pagination at row - The row will be the _last_ row displayed on a
  given page, with a page break after it (or in the future, after any
  footnotes/footer material)

Row types example with 15 lines maximum per page:
(R) indicates a repeated row
(P) indicates a pagination row

```
                 [[1]]
Header                                            ARM1                          ARM2         
Header                                      M              F              F              M   
Header           ----------------------------------------------------------------------------
Content Row      Overall (N)             (N=241)        (N=265)        (N=257)        (N=237)
Label Row        Ethnicity
Content Row        WHITE (n)             (N=123)        (N=129)        (N=133)        (N=114)
Label Row          Factor2
Content Row          Level A             (N=48)         (N=42)         (N=44)         (N=36) 
Label Row            Age Analysis
Data Row               mean               51.46          52.99          55.76          54.32 
Data Row               median             48.15          51.25          56.53          52.1  
Content Row          Level B             (N=40)         (N=43)         (N=38)         (N=49) 
Label Row            Age Analysis
Data Row               mean               55.33          54.47          53.47          55.01 
(P) Data Row           median             56.64          55.13          52.24          54.88 
                 
                 [[2]]
 Header                                            ARM1                          ARM2         
 Header                                      M              F              F              M   
 Header           ----------------------------------------------------------------------------
(R) Content Row  Overall (N)             (N=241)        (N=265)        (N=257)        (N=237)
(R) Label Row    Ethnicity
Content Row         BLACK (n)             (N=118)        (N=136)        (N=124)        (N=123)
Label Row           Factor2
Content Row           Level B             (N=47)         (N=42)         (N=46)         (N=42) 
Label Row             Age Analysis
Data Row                mean               55.84          53.93          56.35          58.52 
Data Row                median             55.29          55.58          55.84          59.9  
Content Row           Level A             (N=42)         (N=47)         (N=47)         (N=37) 
Label Row             Age Analysis
Data Row                mean               56.55          57.03          56.51          54.29 
(P) Data Row            median             56.77          57.56          56.61          54.24 
                 
                 [[3]]
Header                                       ARM1                          ARM2         
Header                                 M              F              F              M   
Header           -----------------------------------------------------------------------
Label Row        Var3 Counts
Data Row           level1           (N=124)        (N=126)        (N=131)        (N=117)
Data Row           level2           (N=117)        (N=139)        (N=126)        (N=120)
                                 

```



```
[[1]]
                                                  ARM1                          ARM2         
                                            M              F              F              M   
                 ----------------------------------------------------------------------------
CR (#1)          Overall (N)             (N=241)        (N=265)        (N=257)        (N=237)
LR (#2)          Ethnicity
CR (#3)            WHITE (n)             (N=123)        (N=129)        (N=133)        (N=114)
LR (#4)            Factor2
CR (#5)              Level A             (N=48)         (N=42)         (N=44)         (N=36) 
LR (#6)              Age Analysis
DR (#7)                mean               51.46          52.99          55.76          54.32 
(P) DR (#8)            median             48.15          51.25          56.53          52.1  
                 
[[2]]
                                  ARM1                          ARM2         
                           M              F              F              M   
                 ----------------------------------------------------------------------------
 (R) CR (#1)     Overall (N)             (N=241)        (N=265)        (N=257)        (N=237)
 (R) LR (#2)     Ethnicity
 (R) CR (#3)       WHITE (n)             (N=123)        (N=129)        (N=133)        (N=114)
 (R) LR (#4)       Factor2
 CR (#9 )            Level B             (N=40)         (N=43)         (N=38)         (N=49) 
 LR (#10)            Age Analysis
 DR (#11)              mean               55.33          54.47          53.47          55.01 
 (P) DR (#12)          median             56.64          55.13          52.24          54.88 
                 
[[3]]
                                                  ARM1                          ARM2         
                                            M              F              F              M   
                 ----------------------------------------------------------------------------
(R) CR (#1)      Overall (N)             (N=241)        (N=265)        (N=257)        (N=237)
(R) LR (#2)      Ethnicity
CR (#13)           BLACK (n)             (N=118)        (N=136)        (N=124)        (N=123)
LR (#14)           Factor2
                     Level B             (N=47)         (N=42)         (N=46)         (N=42) 
                     Age Analysis
                       mean               55.84          53.93          56.35          58.52 
                       median             55.29          55.58          55.84          59.9  
                 
[[4]]
                                 ARM1                          ARM2         
                           M              F              F              M   
----------------------------------------------------------------------------
Overall (N)             (N=241)        (N=265)        (N=257)        (N=237)
Ethnicity
  BLACK (n)             (N=118)        (N=136)        (N=124)        (N=123)
  Factor2
    Level A             (N=42)         (N=47)         (N=47)         (N=37) 
    Age Analysis
      mean               56.55          57.03          56.51          54.29 
      median             56.77          57.56          56.61          54.24 

[[5]]
                            ARM1                          ARM2         
                      M              F              F              M   
-----------------------------------------------------------------------
Var3 Counts
  level1           (N=124)        (N=126)        (N=131)        (N=117)
  level2           (N=117)        (N=139)        (N=126)        (N=120)





```

## Pagination Rules

Pagination is defined to happen at (after) a row if that row will be
the _last_ row on a page, with a page break directly after it (or
after any footnotes which currently are not modeled, but will
be). This is mostly for convenience when coding and could be changed
if it ends up being problematic.

Pagination is performed with respect to a customizable but fixed
maximum number of lines per page (`lpp`). All lines are assumed to be
the same height.

For a given iteration, pagination will occur at the latest row
position (`pos`) such that the following conditions are met:

1. The header lines, repeated rows, and rows between `pos` (inclusive)
and the last page break (non inclusive) together require no more than
`lpp` lines
2. pos points to a Data Row (i.e., NOT a Label or Content Row)
3. _Either_ the Data Row at `pos` does not have any Data Row siblings,
OR pos does not point to the first Data Row data that position in the
tree structure


### The Algorithm.

The algorithm in pseudo code

```

ok_pag_position = functoin(rows, pos, lastpag, reprows, lpp, nhlines) {
     ## not content row
     !is_contentrow(rows[[pos]]) &&
     	 ## not label row
         !is_labelrow(rows[[pos]]) &&
	 ## all required lines fit on page
	 (pos - lastpag + length(reprows) + nhlines) <= lpp &&
	 ## only data row at that tree position, or not first one
	 (is_datarow(rows[[pos - 1]]) || !is_datarow(rows[[pos + 1]]))
}
 	 
get_next_paginate = function(allrows, lpp, lastpag, nhlines) 
	pos = lastpag + lpp - nhlines
	reprows = getreprows(allrows, lastpag)
	while(pos > lastpag && !ok_pag_position(allrows, pos, lastpag, reprows, lpp, nhlines)) {
	     pos = pos - 1
	}
	if(pos == lastpag) stop("unable to find valid pagination positionb between rows", lastpag, "and", lastpag+lpp)
        list(pos = pos, reprows = reprows)
}
```
