---
title: "Titles, Footers, and Referential Footnotes"
author: "Gabriel Becker and Adrian Waddell"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Titles, Footers, and Referential Footnotes}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---


```{r, echo=FALSE}
knitr::opts_chunk$set(comment = "#")
```

```{css, echo=FALSE}
.reveal .r code {
    white-space: pre;
}
```


# Titles and Non-Referential Footer Materials

An `rtables` table can be annotated with three types of header (title)
information, as well as three types of footer information.

Header information comes in two forms that are specified directly
(main title and subtitles), as well as one that is populated automatically
as necessary (page title).

Similarly, footer materials come with two directly specified components:
main footer and provenance footer, in addition to one that is computed
when necessary: referential footnotes.

`basic_table()` accepts the values for each element that can be set directly:

```{r}
library(rtables)
library(dplyr)
lyt <- basic_table(title = "Study XXXXXXXX",
                   subtitles = c("subtitle YYYYYYYYYY", "subtitle2 ZZZZZZZZZ"),
                   main_footer = "Analysis was done using cool methods that are correct",
                   prov_footer = "file: /path/to/stuff/that/lives/there HASH:1ac41b242a") %>%
    split_cols_by("ARM") %>%
    split_rows_by("SEX", page_by = TRUE, page_prefix = "Patient Subset - Gender", split_fun = drop_split_levels) %>%
    split_rows_by("STRATA1") %>%
    analyze("AGE", mean, format = "xx.x")

tbl <- build_table(lyt, DM)
cat(export_as_txt(tbl, paginate = TRUE, page_break = "\n\n\n"))
```
                   


# Referential Footnotes

Referential footnotes are footnotes associated with a particular
component of a table: a column, a row, or a cell. They can be added
during tabulation via analysis functions, but they can also be added
post-hoc once a table is created.

They are rendered as a number within curly braces within the 
table body, row, or column labels, followed by a message associated
with that number printed below the table during rendering.

## Adding Cell- and Analysis-row Referential Footnotes At Tabulation Time

```{r}


afun <- function(df, .var, .spl_context) {
    val <- .spl_context$value[NROW(.spl_context)]
    rw_fnotes <-  if(val == "C") list("This is strata level C for these patients") else list()
    cl_fnotes <- if(val == "B" && df[1, "ARM", drop = TRUE] == "C: Combination")
                     list("these Strata B patients got the drug combination") else list()
    
    in_rows(mean = mean(df[[.var]]),
            .row_footnotes = rw_fnotes,
            .cell_footnotes = cl_fnotes,
            .formats = c(mean = "xx.x"))
}

lyt <- basic_table(title = "Study XXXXXXXX",
                   subtitles = c("subtitle YYYYYYYYYY", "subtitle2 ZZZZZZZZZ"),
                   main_footer = "Analysis was done using cool methods that are correct",
                   prov_footer = "file: /path/to/stuff/that/lives/there HASH:1ac41b242a") %>%
    split_cols_by("ARM") %>%
    split_rows_by("SEX", page_by = TRUE, page_prefix = "Patient Subset - Gender", split_fun = drop_split_levels) %>%
    split_rows_by("STRATA1") %>%
    analyze("AGE", afun, format = "xx.x")

tbl <- build_table(lyt, DM)
cat(export_as_txt(tbl, paginate = TRUE, page_break = "\n\n\n"))

```

We note that typically the type of footnote added within the analysis function would be dependent
on the computations done to calculate the cell value(s), e.g., a model not converging. We simply
use context information as an illustrative proxy for that.

The procedure for adding footnotes to content (summary row) rows or cells
is identical to the above, when done within a content function.

## Annotating an Existing Table with Referential Footnotes

In addition to inserting referential footnotes at tabulation time within
our analysis functions, we cal also annotate our tables with them
post-hoc.

This is also the only way to add footnotes to **column** labels,
as those cannot be controlled within an analysis or content
function.

```{r}

lyt2 <- basic_table() %>%
  split_cols_by("ARM") %>%
    split_cols_by("SEX", split_fun = drop_split_levels) %>%
    split_rows_by("RACE", split_fun = drop_split_levels) %>%
    summarize_row_groups() %>%
    analyze(c("AGE", "STRATA1"))

tbl2 <- build_table(lyt2, ex_adsl %>% filter(SEX %in% c("M", "F") &
                                             RACE %in% (levels(RACE)[1:3])))
tbl2
```


We do this with the `fnotes_at_path<-` function which accepts a row
path, a column path, and a value for *the full set of footnotes for
the defined locations (`NULL` or a `character` vector).

A non-`NULL` row path with a `NULL` column path specifies the footnote(s)
should be attached to the row, while `NULL` row path with non-`NULL`
column path indicates they go with the column. Both being non-null
indicates a cell (and must resolve to an individual cell).

```{r}

fnotes_at_path(tbl2, c("RACE", "ASIAN")) <- c("hi", "there")
tbl2
```

```{r}

fnotes_at_path(tbl2, rowpath = NULL, c("ARM", "B: Placebo")) <- c("this is a placebo")
tbl2
```

Note to step into the content table row we must add that to the path, even though we didn't
need it to put a footnote on the full row.

```{r}
fnotes_at_path(tbl2, rowpath = c("RACE", "ASIAN", "@content", "ASIAN"),  c("ARM", "B: Placebo", "SEX", "F")) <- "These asian women got placebo treatments"
tbl2
```

